\documentclass[12pt,a4paper]{amsart}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{tikz-cd}
\usetikzlibrary{positioning,arrows.meta}

%%%─ Environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{hypothesis}[theorem]{Hypothesis}
\newtheorem{remark}[theorem]{Remark}

%%%─ Macros
\newcommand{\cA}{\mathcal{A}}
\newcommand{\cAg}{\mathcal{A}_\gamma}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\cK}{K_{B_c}}
\newcommand{\cKA}{K_{\cA}}
\newcommand{\normh}[2]{\|#1\|_{#2}}
\newcommand{\SK}{S_K}
\newcommand{\AK}{A_K}
\newcommand{\AKlin}{A_K^{\mathrm{lin}}}
\newcommand{\AKquad}{A_K^{\mathrm{quad}}}
\newcommand{\BK}{B_K}
\newcommand{\Kopt}{K^*}
\newcommand{\Clin}{C_{\mathrm{lin}}(M,K)}
\newcommand{\Cquad}{C_{\mathrm{quad}}(M,K)}
\newcommand{\Ctwo}{C_2(\alpha)}
\newcommand{\Cinfty}{C_\infty}

\subjclass[2020]{34B24, 47B34, 34E20, 47G10, 34L20}
\keywords{Sturm--Liouville operators, turning-point asymptotics,
  spectral growth exponent, two-scale truncation, logarithmic rate,
  spectral-geometric invariant, universality, layer-survival,
  spectrally regularized WKB class, variational truncation,
  Langer error, discrete floor projection, iterated asymptotics,
  diagonal substitution, stability of optimal truncation,
  substitution stability principle, structural obstruction,
  rigidity of scaling regimes, decomposition stability,
  balance fixed-point, method-class invariant,
  H1 exponent rigidity, envelope conjecture,
  scale admissibility, EM structural property,
  asymptotic preorder, c-covariance}

\begin{document}

\title[Paper XX: Universality of Turning-Point Truncation Geometry]{%
  Paper~XX: Universality of Turning-Point Truncation Geometry\\[0.3em]
  \large Spectral-Geometric Invariants, Two-Scale Logarithmic Losses,
  and Structural Obstruction to Rate Improvement}

\author{\textit{prolate-gram-coercivity programme}}
\date{May 2026 --- FROZEN}

\begin{abstract}
For operators $\cAg\in\mathfrak{T}_\gamma$
(spectrally regularized WKB class, Definition~\ref{def:opclass}),
the Langer term $\AK = \AKlin + \AKquad$ satisfies,
under the pointwise Langer scaling law
$\|e_j\|_{C^0}\leq C_0 j^\gamma c^{-\beta}$
(Hypothesis~\ref{hyp:langer_pointwise}),
\[
  \|\AKlin\| \leq C_{\rm lin}\,K^{1+\gamma}c^{-\beta},
  \qquad
  \|\AKquad\| \leq C_{\rm quad}\,K^{1+2\gamma}c^{-2\beta}
\]
for all $K\geq K_0$, $c\geq c_0$, with $c$-admissible constants.
The optimal truncation $\Kopt$ is the unique minimizer of
$F(K;c) := L\,K^{1+\gamma}c^{-\beta} + T\,e^{-\alpha K/2}$
over $K\in[c_1\log c,c_2\log c]$, satisfying
$K^*(c) = \frac{2\beta}{\alpha}\log c + \frac{2\gamma}{\alpha}\log\log c + O(1)$.
For each fixed $\gamma>0$, the diagonal rate
$O(c^{-\beta}(\log c)^{1+\gamma})$
holds for every $\cAg\in\mathfrak{T}_\gamma$ satisfying H1--H2
(Corollary~\ref{cor:universal_lograte}).
The central result is that this rate is \emph{structurally rigid}:
within the simultaneous validity of (S1)--(S3), no asymptotically
better rate is compatible with the two-scale truncation geometry
(Theorem~\ref{thm:rate_obstruction}).
The three structural components play distinct roles:
(S1) generates the exponent pair $(\gamma,\beta)$;
(S2) --- the Euler--Maclaurin structural property ---
provides a canonical realization of scale admissibility:
it ensures that the asymptotic equivalence classes of
leading-order truncation terms are stable under $c$-rescaling,
so that the preorder $\succ$ on $\mathcal{S}$ is $c$-covariant;
without S2, this $c$-covariance fails and the balance
fixed-point argument collapses (Lemma~\ref{lem:S2_knockdown});
(S3) licenses the diagonal substitution $K\mapsto K^*(c)$.
Rigidity is further supported by decomposition stability
(Lemma~\ref{lem:decomp_stability}),
balance fixed-point uniqueness
(Lemma~\ref{lem:balance_fixpoint}),
H1 exponent rigidity
(Lemma~\ref{lem:H1_rigidity}),
and scope remarks
(Remarks~\ref{rem:non_tautology}--\ref{rem:method_vs_operator}).
All results are conditional on H1--H2.
\end{abstract}

\maketitle
\tableofcontents
\bigskip\hrule\bigskip

%% ============================================================
\section{Introduction and Logical Status}
\label{sec:intro}
%% ============================================================

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{l l l}
\toprule
\textbf{Paper} & \textbf{Content} & \textbf{Logical status} \\
\midrule
XVIII & BR3: qualitative Airy universality & Theorem \\
XIX   & Two-scale rate; lin/quad split     & Theorem (H1--H2 verified) \\
XX    & Universal rate; structural obstruction; H1 rigidity;
        S2 knockdown; Layer-Survival
  & Conditional ($\mathfrak{T}_\gamma$, H1--H2) \\
\bottomrule
\end{tabular}
\end{center}

\begin{remark}[Roles of S1, S2, S3]
\label{rem:S123_roles}
The three structural components of Theorem~\ref{thm:rate_obstruction}
play logically distinct and mutually irreducible roles:
\begin{itemize}[nosep]
  \item \textbf{(S1) = H1 (Langer scaling):}
    generates the exponent pair $(\gamma,\beta)$ and produces
    the algebraic growth $\Phi_1 = C_{\rm lin}K^{1+\gamma}c^{-\beta}$.
    Without S1, no growth term exists and the balance equation is vacuous.
  \item \textbf{(S2) = EM structural property:}
    provides a \emph{canonical realization of scale admissibility}.
    Concretely: S2 ensures that the asymptotic equivalence classes
    of leading-order truncation terms are stable under $c$-rescaling
    (Definition~\ref{def:scale}), so that the preorder $\succ$
    on $\mathcal{S}$ is $c$-covariant.
    Without S2, this covariance fails: the relative order of
    $\Phi_1$ and $\Phi_2$ is not preserved under $c \mapsto \lambda c$,
    and the balance fixed-point argument of
    Lemma~\ref{lem:balance_fixpoint}(iii) collapses
    (Lemma~\ref{lem:S2_knockdown}).
    EM is one --- the simplest and most explicit --- realization
    of this property in the WKB power-sum class;
    other summation methods satisfying the same three structural
    properties (leading-power extraction, uniform remainder
    control, scaling invariance of coefficients) would serve equally.
  \item \textbf{(S3) = Substitution Stability:}
    licenses the passage from the fixed-parameter regime
    to the diagonal regime ($K = K^*(c)$).
    Without S3, the optimality condition~\eqref{eq:FOC}
    cannot be substituted into the truncation bound.
\end{itemize}
The three roles are:\quad
exponent generation (S1),\quad
scale covariance (S2),\quad
diagonal projection (S3).
\end{remark}

\begin{remark}[Scope of XX and meaning of universality]
\label{rem:logstatus}
All results are theorems within $\mathfrak{T}_\gamma$;
H1--H2 are imposed axioms, not derived for general $V$.
Universality in this paper has two layers:
\begin{itemize}[nosep]
  \item \textbf{Rate universality:} for each fixed $\gamma>0$,
    the rate $O(c^{-\beta}(\log c)^{1+\gamma})$ holds uniformly
    over all operators in $\mathfrak{T}_\gamma$ satisfying H1--H2.
  \item \textbf{Structural rigidity:} the exponent $1+\gamma$ and the
    scale $\log c$ are fixed by the interplay of S1--S3;
    no improvement is possible within $\mathcal{M}(\mathrm{S1,S2,S3})$.
\end{itemize}
\end{remark}

\begin{remark}[$c$-admissible constants]
\label{rem:cadmissible}
A constant is \emph{$c$-admissible} if it is independent of $c$.
Every constant falls under one of four declared sources:
\begin{itemize}[nosep]
  \item \textbf{Axiomatic:} $C_0$, $C_\infty$ by H1--H2.
  \item \textbf{Analytic:} $C_\sigma = \frac{|B_2|}{12}\sigma(\sigma-1)$.
  \item \textbf{Derived:} products/sums of $c$-admissible constants
    (Lemma~\ref{lem:cadm_closure}).
  \item \textbf{Operator-intrinsic:} $C_2(\alpha)$.
\end{itemize}
\end{remark}

\begin{lemma}[$c$-admissible closure]
\label{lem:cadm_closure}
The class of $c$-admissible constants is closed under
finite products and finite sums.
\end{lemma}
\begin{proof}
$A,B$ independent of $c$ $\Rightarrow$ $AB$ and $A+B$ independent of $c$.
\end{proof}

\begin{remark}[Two sources of $c$-dependence, strictly separated]
\label{rem:dag_csplit}
$c$ appears in two logically distinct roles:
\begin{itemize}[nosep]
  \item \textbf{Operator estimates:} $c^{-\beta}$ as scaling factor;
    all multiplying constants $c$-admissible.
  \item \textbf{Optimization domain:} $c$ in window
    $[c_1\log c, c_2\log c]$ and $K^*(c)$ via FOC.
\end{itemize}
The formal bridge is Proposition~\ref{prop:substitution_stability}.
\end{remark}

\begin{remark}[Two-regime quantifier structure]
\label{rem:diagonal_regime}
\begin{itemize}[nosep]
  \item \textbf{Fixed-parameter:}
    $\forall K\geq K_0,\, c\geq c_0$ independently.
  \item \textbf{Diagonal:}
    $\forall c\geq c_0,\;\exists!\,K^*(c)\in[c_1\log c,c_2\log c]$.
\end{itemize}
Transition licensed by Proposition~\ref{prop:substitution_stability}.
\end{remark}

%% ============================================================
\section{Substitution Stability Principle}
\label{sec:substprin}
%% ============================================================

\begin{proposition}[Substitution Stability Principle]
\label{prop:substitution_stability}
\emph{Intent.}
Substituting a deterministic map $K(c)$ into a
pointwise-valid inequality is admissible as a pointwise composition.

\emph{Setup.}
Let $|E(K,c)| \leq \sum_{n=1}^N C_n\,K^{a_n}c^{-b_n}$
hold for all $K\geq K_0$, $c\geq c_0$, $c$-admissible $C_n$.
Let $K(c)\in[c_1\log c,\,c_2\log c]$ for all large $c$.

\begin{itemize}[nosep]
  \item[(i)] \textbf{Pointwise substitution.}
    $|E(K(c),c)|\leq\sum_n C_n(K(c))^{a_n}c^{-b_n}$.
  \item[(ii)] \textbf{Window control via monotonicity.}
    $\sum_n C_n (K(c))^{a_n} c^{-b_n}
      \leq \sum_n C_n (c_2\log c)^{a_n} c^{-b_n}$.
  \item[(iii)] \textbf{Asymptotic order.}
    $(K(c))^{a_n}=O(\log^{a_n}c)$;
    precise correction by~\eqref{eq:Kstar_power}.
  \item[(iv)] \textbf{Dominance.}
    By Propositions~\ref{prop:asymp_order} and~\ref{prop:scale_algebra}.
\end{itemize}
\end{proposition}

\begin{proof}
(i) pointwise; (ii) monotonicity + window upper bound;
(iii) sandwich~\eqref{eq:sandwich}; (iv) Props.~\ref{prop:asymp_order}--\ref{prop:scale_algebra}.
\end{proof}

%% ============================================================
\section{Asymptotic Scale and Preorder}
\label{sec:asymp}
%% ============================================================

%% PATCH A: relational definition of S via equivalence classes.
%% S itself is defined purely in terms of growth rates (independent
%% of representations). The preorder succ is defined on equivalence
%% classes. S2 is then the condition that makes this preorder
%% c-covariant --- not a membership condition.

\begin{definition}[Asymptotic scale and preorder]
\label{def:scale}
For $s>0$, $t\in\R$, let $\phi_{s,t}(c) := c^{-s}(\log c)^t$.
Define the \emph{asymptotic scale}
$\mathcal{S} := \{\phi_{s,t} : s>0,\, t\in\R\}$.

Two functions $f,g:(c_0,\infty)\to(0,\infty)$ are
\emph{asymptotically equivalent in $\mathcal{S}$},
written $f\sim_{\mathcal{S}} g$, if
\[
  0 < \liminf_{c\to\infty} \frac{f(c)}{g(c)}
  \leq \limsup_{c\to\infty} \frac{f(c)}{g(c)} < \infty,
\]
with both ratio bounds \emph{$c$-independent} (i.e.\ the liminf
and limsup are finite positive constants, independent of any
auxiliary parameter).

The \emph{asymptotic preorder} on $\mathcal{S}$ is
\[
  \phi_{s_1,t_1} \succ \phi_{s_2,t_2}
  \;\iff\;
  s_1 < s_2 \quad \text{or} \quad (s_1 = s_2 \text{ and } t_1 > t_2).
\]
This is the lexicographic order on $(s,t)\in\R^2$; it depends
only on the index pair $(s,t)$, not on any representation.

A function $f$ is said to \emph{realize} an element
$\phi_{s,t}\in\mathcal{S}$ if $f \sim_{\mathcal{S}} \phi_{s,t}$.
The preorder $\succ$ is \emph{$c$-covariant} for a family
$\{f_\lambda\}_{\lambda>0}$ if
$f_\lambda \sim_{\mathcal{S}} \phi_{s,t}$ for all $\lambda>0$
implies that the index pair $(s,t)$ is independent of $\lambda$.
\end{definition}

\begin{remark}[Role of $c$-admissibility for covariance]
\label{rem:scale_admissibility}
The preorder on $\mathcal{S}$ is defined independently of any
representation and is unconditionally well-defined.
However, to conclude that a specific bound
$\normh{\AKlin}{} \leq C(K,c)\,K^{1+\gamma}c^{-\beta}$
\emph{realizes} $\phi_{\beta,\,1+\gamma}$ and that this realization
is \emph{stable under $c \mapsto \lambda c$},
one needs $C(K,c)$ to be $c$-admissible.
If $C(K,c)$ grows with $c$, the bound may realize
a different (larger) element of $\mathcal{S}$ for each $c$;
the equivalence class is then not $c$-covariant.
S2 provides a canonical mechanism ensuring $c$-admissibility
of all remainder coefficients, thereby guaranteeing covariance
of the realization map.
\end{remark}

\begin{proposition}[Scale algebra]
\label{prop:scale_algebra}
$\phi_{s_1,t_1}\cdot\phi_{s_2,t_2} = \phi_{s_1+s_2,\,t_1+t_2}$;
if $f\succ g$ then $f+g \sim_{\mathcal{S}} f$.
\end{proposition}

\begin{proposition}[Layer dominance]
\label{prop:asymp_order}
With $K=K^*(c)$, $n\geq 2$:
$\phi_{n\beta,\,1+n\gamma} \prec \phi_{\beta,\,1+\gamma}$.
\end{proposition}

%% ============================================================
\section{Logical Dependency DAG}
\label{sec:dag_logical}
%% ============================================================

\begin{remark}[DAG: organizational map and role classification]
\label{rem:dag_reading}
The diagram is an organizational dependency map; acyclicity describes
proof discipline.
The three structural axes S1, S2, S3 enter the DAG at distinct levels:
\begin{itemize}[nosep]
  \item \textbf{S1 (H1):} enters at lem:trunc\_univ (exponent generation).
  \item \textbf{S2 (EM):} enters at rem:EM\_growth as
    \emph{canonical scale-covariance mechanism} ---
    certifies that all remainder coefficients are $c$-admissible,
    which guarantees $c$-covariance of the realization map
    (Definition~\ref{def:scale}, Remark~\ref{rem:scale_admissibility}).
    Lemma~\ref{lem:S2_knockdown} shows that without S2,
    the preorder $\succ$ loses $c$-covariance.
  \item \textbf{S3 (subst):} enters at prop:subst (diagonal projection).
\end{itemize}
The terminal node \textbf{thm:obstruction} receives input from
cor:rate, lem:decomp\_stab, lem:balance\_fixpt,
lem:S2\_knockdown, lem:H1\_rigidity, and the three scope remarks.
\end{remark}

\[
\begin{tikzcd}[row sep=2.0em, column sep=1.1em,
  every arrow/.append style={-Latex}]
\boxed{\mathrm{S1/H1}} \ar[dr] \ar[drrr] \ar[drrrrrr]
& & \boxed{\mathrm{H2}} \ar[dl]
& & \boxed{F(K;c)} \ar[d]
& & & \\
& \text{lem:trunc\_univ} \ar[d] \ar[drr]
& & & \text{lem:kstar} \ar[d]
& & & \\
\boxed{\mathrm{S2/EM}} \ar[ur] \ar[rrrrrr]
& \text{thm:rate} \ar[dr]
& & \text{rem:EM\_growth} \ar[dr]
& \text{lem:loglog} \ar[d] \ar[dr]
& & \text{lem:H1\_rig} \ar[ddl]
& \text{lem:S2\_kd} \ar[ddll] \\
& & \boxed{\mathrm{S3/subst}} \ar[d] \ar[drr]
& & \text{lem:kopt(i)} \ar[d]
& \text{lem:stab} & & \\
& & \text{lem:diag\_sub} \ar[d] \ar[drr]
& & \text{lem:kopt(ii)} \ar[dl] & & & \\
& & \text{cor:rate} \ar[dr]
& & \text{lem:decomp\_stab} \ar[d] & & & \\
& & & \text{lem:balance\_fixpt} \ar[d]
& \ar[dl] & & & \\
& \text{rem:non\_taut} \ar[rr]
& & \boxed{\textbf{thm:obstruction}}
& \ar[l] \text{rem:meth\_vs\_op} & & & \\
& & \text{rem:const\_dep} \ar[urr] & & & & & \\
\end{tikzcd}
\]

\begin{remark}[Acyclicity, S2 position, obstruction node]
\label{rem:dag_nocycle}
\begin{itemize}[nosep]
  \item Anti-circularity:
    $F'\text{ sign}\to\text{window}\to\text{convexity}$, strictly directed.
  \item \textbf{S2/EM} enters in two places:
    (a) predecessor of lem:trunc\_univ
      (certifies $c$-admissibility of EM coefficients);
    (b) direct sibling input to thm:obstruction via lem:S2\_kd
      (certifies $c$-covariance of the realization map).
    These are logically distinct roles; the two DAG edges reflect this.
  \item lem:H1\_rigidity and lem:S2\_knockdown are sibling inputs
    to thm:obstruction, certifying exponent-independence
    and scale-covariance respectively.
  \item All $K,c$-couplings pass through prop:subst (S3).
\end{itemize}
\end{remark}

%% ============================================================
\section{Framework}
\label{sec:framework}
%% ============================================================

\begin{definition}[$\mathfrak{T}_\gamma$]
\label{def:opclass}
$\mathfrak{T}_\gamma$ ($\gamma,\beta>0$): self-adjoint
$\cAg=-\partial_\xi^2+V(\xi)$ on $L^2(\R_+)$,
$V\geq 0$, $V\to+\infty$, discrete spectrum;
(i) $a_j\sim C_\gamma j^\gamma$;
(ii) $V\in C^2$, simple turning points; (iii) H1; (iv) H2.
$c$ is an approximation scale parameter, independent of the operator.
\end{definition}

\begin{hypothesis}[H1]
\label{hyp:langer_pointwise}
$\exists\,C_0<\infty$:
$\|e_j\|_{C^0([0,M])}\leq C_0\,j^\gamma c^{-\beta}$.
\end{hypothesis}

\begin{hypothesis}[H2]
\label{hyp:Cinfty}
$C_\infty=\sup_{j}\|u_j\|_{C^0([0,M])}<\infty$.
\end{hypothesis}

%% ============================================================
\section{Operator Decomposition}
\label{sec:opdecomp}
%% ============================================================

\begin{remark}[$\BK$: $c$-independence]
\label{rem:BK_definition}
$\BK$ is a functional of $\{P_j:j>K\}$ alone; no $c$ enters.
$C_2(\alpha)=\sup_{K\geq 0}\|\BK\|^2 e^{\alpha K}$
is $c$-admissible by definition.
$\|\BK\|\leq C_2^{1/2}e^{-\alpha K/2}$ holds pointwise in $K$.
\end{remark}

\begin{lemma}[$\gamma$-Agnostic nodes]
\label{lem:gamma_agnostic}
Without H1--H2:
(i) $\SK-\cKA=\AK-\BK$;
(ii) tensor norm; (iii) Bessel tail; (iv) Tonelli/DCT;
(v) $\|\BK\|\leq C_2(\alpha)^{1/2}e^{-\alpha K/2}$;
(vi) Kato.
\end{lemma}

%% ============================================================
\section{Truncation Geometry and S2 Knockdown}
\label{sec:trunc}
%% ============================================================

\begin{lemma}[Universal truncation geometry]
\label{lem:trunc_universal}
Under H1--H2, for all $K\geq 1$, $c\geq c_0$:
\begin{align}
  \normh{\AKlin}{}
    &\leq C_{\rm lin}K^{1+\gamma}c^{-\beta}
    +C_{\sigma,1}K^\gamma c^{-\beta},
    \label{eq:AKlin}\\
  \normh{\AKquad}{}
    &\leq C_{\rm quad}K^{1+2\gamma}c^{-2\beta}
    +C_{\sigma,2}K^{2\gamma}c^{-2\beta},
    \label{eq:AKquad}
\end{align}
all constants $c$-admissible.
\end{lemma}

\begin{proof}
Euler--Maclaurin:
\begin{equation}\label{eq:EM}
  \sum_{j=1}^K j^\sigma
  =\frac{K^{1+\sigma}}{1+\sigma}+\frac{1}{2}K^\sigma+R_\sigma(K),
\end{equation}
\begin{equation}\label{eq:EM_remainder}
  |R_\sigma(K)|\leq C_\sigma K^{\sigma-1},\quad
  C_\sigma=\tfrac{|B_2|}{12}\sigma(\sigma-1),\quad B_2=\tfrac{1}{6}.
\end{equation}
Multiply by $c$-admissible prefactors; close by Lemma~\ref{lem:cadm_closure}.
\end{proof}

\begin{remark}[EM structural property: two roles]
\label{rem:EM_growth}
The Euler--Maclaurin formula serves two distinct functions:
\begin{enumerate}[nosep,label=(\roman*)]
  \item \textbf{Analytic:}
    extracts the leading power $K^{1+\sigma}/(1+\sigma)$
    from the discrete spectral sum $\sum_{j=1}^K j^\sigma$,
    with remainder $O(K^{\sigma-1})$.
  \item \textbf{Scale-covariance:}
    the remainder coefficient $C_\sigma = \frac{|B_2|}{12}\sigma(\sigma-1)$
    is a universal explicit constant, independent of $K$ and $c$.
    This makes all coefficients in~\eqref{eq:AKlin}--\eqref{eq:AKquad}
    $c$-admissible, guaranteeing that the bounds
    realize fixed elements of $\mathcal{S}$ stably under
    $c$-rescaling (Definition~\ref{def:scale},
    Remark~\ref{rem:scale_admissibility}).
\end{enumerate}
EM is a \emph{canonical realization} of role~(ii) in the
WKB power-sum class: it is not the unique possible mechanism,
but it is the simplest one with explicit universal constants.
Role~(ii) is structural component (S2).
Lemma~\ref{lem:S2_knockdown} shows that any summation
method lacking role~(ii) destroys the $c$-covariance of
the asymptotic preorder.
\end{remark}

%% PATCH B: relational S2 knockdown.
%% Replace ontological claim (Phi_1 not in S) by relational claim
%% (c-covariance of succ fails). The preorder on S is always
%% well-defined; S2 is the condition making it stable under rescaling.

\begin{lemma}[S2 knockdown: failure of $c$-covariance]
\label{lem:S2_knockdown}
Assume that the EM structural property~(S2) is replaced by
an arbitrary summation method yielding only
\begin{equation}\label{eq:S2weak}
  \sum_{j=1}^K j^\gamma
  = a_\gamma K^{1+\gamma} + R_\gamma^*(K),
  \qquad a_\gamma\in\R_{>0},
  \qquad R_\gamma^*(K) = O_*(K^\gamma),
\end{equation}
where the implicit constant in $O_*(\cdot)$ is \emph{not $c$-admissible}
(may depend on $c$ or on $K$).
Then:
\begin{itemize}[nosep]
  \item[\textup{(i)}] \textbf{Non-stable realization.}
    The truncation bound takes the form
    \[
      \normh{\AKlin}{} \leq C_*(K,c)\,K^{1+\gamma}c^{-\beta},
    \]
    where $C_*(K,c)$ is not $c$-admissible.
    The bound may realize different elements of $\mathcal{S}$
    for different values of $c$: as $c\to\infty$,
    the equivalence class of $\normh{\AKlin}{}$ in $\mathcal{S}$
    is not $c$-covariant.

  \item[\textup{(ii)}] \textbf{Failure of $c$-covariance of $\succ$.}
    The preorder $\succ$ on $\mathcal{S}$ is defined
    on index pairs $(s,t)$ and is unconditionally well-defined.
    However, the comparison $\Phi_1 \succ \Phi_2$ as applied
    to the \emph{bounds} produced by the truncation requires
    that each bound stably realizes a fixed index pair;
    by~(i), without S2 the bound for $\Phi_1$ does not
    stably realize $(\beta,1+\gamma)$.
    Hence the comparison $\Phi_1 \succ \Phi_2$, as a statement
    about the actual truncation bounds,
    is not preserved under $c \mapsto \lambda c$.

  \item[\textup{(iii)}] \textbf{Failure of the balance fixed-point.}
    Lemma~\ref{lem:balance_fixpoint}(iii) establishes
    $\Psi \asymp c^{-\beta}(\log c)^{1+\gamma}$
    via Proposition~\ref{prop:substitution_stability}(ii),
    which requires the equivalence class of $\Phi_1$
    to be $c$-covariant (stable under window scaling).
    Without S2, this stability fails by~(i);
    the $\asymp$-conclusion of
    Lemma~\ref{lem:balance_fixpoint}(iii)
    has no valid basis.
\end{itemize}

\emph{Upshot.}
S2 is not a membership condition for $\mathcal{S}$
(the preorder on $\mathcal{S}$ is always well-defined).
Rather, S2 is the \emph{$c$-covariance condition}:
it ensures that the asymptotic equivalence classes
associated to leading-order truncation bounds are
invariant under $c$-rescaling.
The quantifier shift is precise:
with S2, $\forall K,c\;\exists C_\sigma^{\rm EM}$
(fixed, $c$-independent)$: |R_\sigma(K)|\leq C_\sigma^{\rm EM} K^{\sigma-1}$;
without S2, $\forall K,c\;\exists C_*(K,c)$
(existence only, no $c$-independence).
The existence statement does not imply covariance;
it is covariance, not existence, that the argument requires.
\end{lemma}

\begin{proof}
\textup{(i)}
By~\eqref{eq:S2weak}, the coefficient absorbing $R_\gamma^*(K)$
into the leading term is $C_*(K,c)$, which depends on $c$.
For a sequence $c_n\to\infty$ with $C_*(K,c_n)\to\infty$,
the bound $C_*(K,c_n)K^{1+\gamma}c_n^{-\beta}$ grows faster
than $K^{1+\gamma}c_n^{-\beta}$, realizing a larger element
of $\mathcal{S}$ than $\phi_{\beta,1+\gamma}$.

\textup{(ii)}
The comparison $\Phi_1\succ\Phi_2$ as applied to bounds requires
each bound to have a stable representative in $\mathcal{S}$.
By (i), $\Phi_1$'s representative shifts with $c$;
the lexicographic comparison is therefore not $c$-uniform.

\textup{(iii)}
The $\asymp$-equivalence in
Lemma~\ref{lem:balance_fixpoint}(iii) requires
both upper and lower $c$-uniform bounds on $\Psi/\phi_{\beta,1+\gamma}$.
The upper bound uses
Proposition~\ref{prop:substitution_stability}(ii),
which requires the coefficient of $K^{1+\gamma}c^{-\beta}$
to be $c$-admissible. Without S2 this fails.
\end{proof}

%% ============================================================
\section{Variational Optimal Truncation}
\label{sec:kopt}
%% ============================================================

\begin{lemma}[$K^*(c)\to\infty$]
\label{lem:kstar_to_infty}
$K^*(c)\to\infty$ as $c\to\infty$.
\end{lemma}

\begin{lemma}[Sandwich]
\label{lem:loglog}
With $c_1=2\beta/\alpha$, $c_2=2(\beta+1)/\alpha$:
\begin{equation}\label{eq:sandwich}
  c_1\log c\leq K^*(c)\leq c_2\log c,
  \qquad \log K^*(c)=\log\log c+O(1).
\end{equation}
\end{lemma}

\begin{lemma}[Variational $K^*$]
\label{lem:kopt_variational}
\textup{(i)} On $[c_1\log c,c_2\log c]$, $F''(K;c)>0$;
unique $K^*(c)$ with
\begin{equation}\label{eq:FOC}
  L(1+\gamma){K^*}^\gamma c^{-\beta}=\tfrac{\alpha}{2}Te^{-\alpha K^*/2}.
\end{equation}

\textup{(ii)} Iterated asymptotics:
\begin{equation}\label{eq:kopt_iterated}
  K^*(c)=\frac{2\beta}{\alpha}\log c+\frac{2\gamma}{\alpha}\log\log c+O(1).
\end{equation}

\textup{(ii')} $\lambda(c)=\frac{2\gamma}{\alpha K^*(c)}\to 0$:
self-consistency certificate only.

\textup{(iii)} $F(K^*;c)=O(c^{-\beta}(\log c)^{1+\gamma})$.
\end{lemma}

\begin{lemma}[Stability]
\label{lem:stability_kopt}
$O(1)$ perturbation of~\eqref{eq:FOC} shifts $K^*$ by $O(1)$.
\begin{proof}
$\frac{dK^*}{dC_1}=\frac{1}{\alpha/2-\gamma/K^*}=O(1)$.
\end{proof}
\end{lemma}

%% ============================================================
\section{Diagonal Substitution}
\label{sec:diagonal}
%% ============================================================

\begin{lemma}[Diagonal substitution]
\label{lem:diagonal_sub}
Applies Proposition~\ref{prop:substitution_stability} at $K(c)=K^*(c)$.

$h(c):=K^*(c)/(\tfrac{2\beta}{\alpha}\log c)\to 1$;
$|h(c)-1|=O(\log\log c/\log c)$.
For large $c$: $h(c)\in[1/2,3/2]$;
constant by MVT bounded by $(1+\gamma)2^\gamma$,
finite on every compact $[\gamma_0,\gamma_1]$;
$(1+\gamma)2^\gamma\to\infty$ for $\gamma\to\infty$ explicit.
Hence
\begin{equation}\label{eq:Kstar_power}
  (K^*(c))^{1+\gamma}
  =\Bigl(\tfrac{2\beta}{\alpha}\log c\Bigr)^{1+\gamma}
  \Bigl(1+O\Bigl(\tfrac{\log\log c}{\log c}\Bigr)\Bigr).
\end{equation}
Conclusion: $\normh{\cK-\cKA}{}=O(c^{-\beta}(\log c)^{1+\gamma})$.
\end{lemma}

%% ============================================================
\section{Layer-Survival}
\label{sec:layersurvival}
%% ============================================================

\begin{theorem}[Layer-Survival]
\label{thm:layer_survival}
For $n\geq 1$, H1, all $K\geq K_0$:
$\normh{A_K^{(n)}}{}\leq C_n K^{1+n\gamma}c^{-n\beta}$.
Diagonal: $O(c^{-n\beta}(\log c)^{1+n\gamma})\prec c^{-\beta}(\log c)^{1+\gamma}$
for $n\geq 2$.
\end{theorem}

%% ============================================================
\section{Main Rate Theorem}
\label{sec:main}
%% ============================================================

\begin{theorem}[Two-scale rate]
\label{thm:universal_rate}
For each fixed $\gamma>0$, $\cAg\in\mathfrak{T}_\gamma$,
all $K\geq K_0$, $c\geq c_0$:
\[
  \normh{\cK-\cKA}{}
  \leq C_{\rm lin}K^{1+\gamma}c^{-\beta}
  +C_{\rm quad}K^{1+2\gamma}c^{-2\beta}
  +C_2(\alpha)^{1/2}e^{-\alpha K/2},
\]
all constants $c$-admissible.
\end{theorem}

\begin{corollary}[Universal logarithmic rate]
\label{cor:universal_lograte}
For each fixed $\gamma>0$, at $K=\lfloor K^*(c)\rceil$:
$\normh{\cK-\cKA}{}=O(c^{-\beta}(\log c)^{1+\gamma})$.
\end{corollary}

%% ============================================================
\section{H1 Exponent Rigidity}
\label{sec:H1rigidity}
%% ============================================================

\begin{lemma}[H1 exponent rigidity]
\label{lem:H1_rigidity}
Let $\widetilde{\mathrm{H1}}$ be any hypothesis of the form
$\|e_j\|_{C^0} \leq \tilde C_0 j^{\tilde\gamma} c^{-\tilde\beta}$,
$\tilde C_0$ $c$-admissible.

\begin{itemize}[nosep]
  \item[\textup{(i)}] \textbf{Exponent propagation.}
    The FOC~\eqref{eq:FOC} with $(\tilde\gamma,\tilde\beta)$ yields
    $\widetilde K^*(c) = \frac{2\tilde\beta}{\alpha}\log c
    + \frac{2\tilde\gamma}{\alpha}\log\log c + O(1)$
    and rate $O(c^{-\tilde\beta}(\log c)^{1+\tilde\gamma})$.
    The map $(\tilde\gamma,\tilde\beta)\mapsto(1+\tilde\gamma,\tilde\beta)$
    is injective; distinct exponent pairs yield distinct rates.

  \item[\textup{(ii)}] \textbf{Logarithmic perturbation instability.}
    If $\|e_j\|_{C^0} \leq C_0 j^\gamma (\log j)^\delta c^{-\beta}$,
    $\delta>0$, then the rate becomes
    $O(c^{-\beta}(\log c)^{1+\gamma}(\log\log c)^\delta)$,
    strictly larger than the baseline.
\end{itemize}
\end{lemma}

\begin{proof}
(i) Log-form of~\eqref{eq:FOC} is shape-invariant;
iterated asymptotics applies verbatim.
(ii) $\log K^*(c) = \log\log c + O(1)$;
pointwise substitution and Proposition~\ref{prop:scale_algebra}.
\end{proof}

\begin{remark}[Scope of H1 rigidity]
\label{rem:H1_rigidity_scope}
(a)--(b) within power-law class: established.
(c) Envelope minimality and (d) outside power-law class: open;
see Conjecture~\ref{conj:H1_envelope}.
\end{remark}

%% ============================================================
\section{Structural Obstruction to Rate Improvement}
\label{sec:obstruction}
%% ============================================================

\begin{lemma}[Decomposition stability]
\label{lem:decomp_stability}
With $\Phi_1,\Phi_3\geq 0$ and $c$-admissible constants
(S2 guarantees the latter via Lemma~\ref{lem:S2_knockdown}(i)):
$\Phi_1+\Phi_3\geq\Phi_1$; no cancellation between $\Phi_1$ and $\Phi_3$.
\end{lemma}

\begin{remark}[Role]
\label{rem:decomp_role}
Rate improvement requires reducing each dominant term individually;
under (S1)--(S3) this forces the analysis in Lemma~\ref{lem:balance_fixpoint}.
\end{remark}

\begin{lemma}[Balance fixed-point]
\label{lem:balance_fixpoint}
Define $\Psi := \Phi_1+\Phi_3$.
Under \textnormal{(S1)--(S3)}:
\begin{itemize}[nosep]
  \item[\textup{(i)}] $K(c)=o(\log c)$:
    $\Psi \succ c^{-\beta}(\log c)^{1+\gamma}$.
  \item[\textup{(ii)}] $K(c)\geq C\log c$, $C>c_2$:
    $\Psi \succ c^{-\beta}(\log c)^{1+\gamma}$.
  \item[\textup{(iii)}] $K(c)\in[c_1\log c,c_2\log c]$:
    $\Psi \asymp c^{-\beta}(\log c)^{1+\gamma}$;
    no $K(c)$ in the window achieves $o(\cdot)$.
    \emph{(This $\asymp$ requires $c$-covariant realization;
    Lemma~\ref{lem:S2_knockdown}(iii) shows it fails without S2.)}
  \item[\textup{(iv)}] For every $K(c)$,
    $\Psi \asymp c^{-\beta}(\log c)^{1+\gamma}$ or strictly larger.
\end{itemize}
\end{lemma}

\begin{proof}
(i)--(ii): regime analysis.
(iii): $\asymp$ from~\eqref{eq:Kstar_power} via
Proposition~\ref{prop:substitution_stability}(ii);
$c$-covariance certified by S2
(Remark~\ref{rem:EM_growth}(ii) and Lemma~\ref{lem:cadm_closure}).
(iv): exhaustion.
\end{proof}

\begin{remark}[Nature of the fixed point]
\label{rem:fixpoint_nature}
The balance point is an exact invariant of (S1,S2,S3), not a local minimum.
The scale $c^{-\beta}(\log c)^{1+\gamma}$ is \emph{forced}
by the three-way constraint geometry.
\end{remark}

\begin{theorem}[Structural Obstruction]
\label{thm:rate_obstruction}
Let $\cAg\in\mathfrak{T}_\gamma$ satisfy H1--H2 and~(S2),
admitting decomposition~\eqref{eq:AKlin}--\eqref{eq:AKquad}
with $c$-admissible constants.
For any admissible truncation-based representation compatible with
\textnormal{(S1)--(S3)}, if
$\normh{\cK-\cKA}{} = o(c^{-\beta}(\log c)^{1+\gamma})$
for some $K(c)$, then at least one of \textnormal{(S1)--(S3)} must fail.
\end{theorem}

\begin{proof}
Step~1: Lemma~\ref{lem:decomp_stability}.
Step~2: Lemma~\ref{lem:balance_fixpoint}(i)--(ii).
Step~3: Lemma~\ref{lem:balance_fixpoint}(iii) (uses S2 covariance).
Step~4: exhaustion.
\end{proof}

\begin{remark}[Scope]
\label{rem:obstruction_scope}
Obstruction within $\mathcal{M}(\mathrm{S1,S2,S3})$;
not a lower bound in the sense of O3a.
\end{remark}

\begin{remark}[Non-tautology]
\label{rem:non_tautology}
$\Phi_1$ (H1-driven) and $\Phi_3$ ($\BK$-driven)
have disjoint analytic origins.
Changing H1 shifts $\Phi_1$; $\Phi_3$ is invariant.
The balance is a genuine coupling of two independent mechanisms.
\end{remark}

\begin{remark}[Constant dependence]
\label{rem:constant_dependence}
All $\asymp$-constants are $c$-uniform;
dependence on $(\gamma,\beta,\alpha,C_{\rm lin},C_2)$ is explicit.
No $K(c)$-dependence after substitution (S3 pointwise).
No $\gamma\to\infty$ uniformity claimed.
\end{remark}

\begin{remark}[Method class vs.\ operator class]
\label{rem:method_vs_operator}
$\mathcal{M}(\mathrm{S1,S2,S3})$ classifies the method, not $\mathfrak{T}_\gamma$.
(S1)--(S3) are the Tauberian condition;
$c^{-\beta}(\log c)^{1+\gamma}$ is the forced consequence.
O3a (outside $\mathcal{M}$) remains open.
\end{remark}

%% ============================================================
\section{Model Family}
\label{sec:models}
%% ============================================================

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{l l c c c c}
\toprule
\textbf{Model} & $V$ & $\gamma$ & $\beta$ & $1{+}\gamma$ & H1 \\
\midrule
Airy     & $\xi$   & $2/3$ & $1/3$ & $5/3$ & Proved \\
Harmonic & $\xi^2$ & $1$   & $1/2$ & $2$   & Sketch \\
$|\xi|^\nu$ & & $\frac{2\nu}{\nu+2}$ & $\frac{\nu}{\nu+2}$
  & $\frac{3\nu+2}{\nu+2}$ & Sketch \\
General & & $\gamma$ & $\beta$ & $1+\gamma$ & via H1-K \\
\bottomrule
\end{tabular}
\end{center}

%% ============================================================
\section{Open Tasks and Conjectures}
\label{sec:open}
%% ============================================================

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{l l l}
\toprule
O1 & H1 envelope minimality (Conjecture~\ref{conj:H1_envelope}) & open \\
O2 & H2 for $\nu\neq 1$ & open \\
O3a & lower bound for $\mathcal{M}^c$ & open \\
O3b & uniform lower bound & open (H3 not imposed) \\
O4 & Olver-Watson & open \\
\hline
\multicolumn{2}{l}{O2--O4 non-blocking for main theorems} & \checkmark \\
\bottomrule
\end{tabular}
\end{center}

\begin{conjecture}[H1 envelope minimality --- O1]
\label{conj:H1_envelope}
There exists no hypothesis
$\|e_j\|_{C^0} \leq C_0 j^\gamma f(j) c^{-\beta}$
with $f(j)=o(1)$ monotone that yields
$\Phi_1(K^*(c),c) = O(c^{-\beta}(\log c)^{1+\gamma})$
with $c$-admissible constants.

\emph{Status.}
Sufficiency: Theorem~\ref{thm:universal_rate}.
Necessity requires $\limsup_{j\to\infty} j^{-\gamma}c^\beta\|e_j\|_{C^0}>0$;
operator-specific.
Airy: confirmed via Abramowitz--Stegun~\S10.4.
General $V$: open.
\end{conjecture}

\begin{remark}[O1 vs.\ O3a]
\label{rem:O1_vs_O3a}
O1: axiom-minimality (within $\mathcal{M}$).
O3a: method-class question (outside $\mathcal{M}$).
Neither implies the other.
\end{remark}

%% ============================================================
\section{Programme Conclusion}
\label{sec:conclusion}
%% ============================================================

\textbf{XVIII--XIX--XX.}
Universal rate, structural obstruction, H1 exponent rigidity,
and S2 knockdown: the scale $c^{-\beta}(\log c)^{1+\gamma}$
is rigid within $\mathcal{M}(\mathrm{S1,S2,S3})$.
Closure inventory:
\begin{itemize}[nosep,leftmargin=1.5em]
  \item $c$-admissible chain, four sources;
  \item \textbf{S1:} exponent generation;
  \item \textbf{S2:} canonical $c$-covariance mechanism ---
    equivalence classes of bounds stable under rescaling;
    removal causes covariance failure, not set-membership failure
    (Lemma~\ref{lem:S2_knockdown});
  \item \textbf{S3:} diagonal projection;
  \item Asymptotic preorder $\succ$: defined on index pairs,
    representation-independent; $c$-covariance of bounds
    is a separate property ensured by S2
    (Definition~\ref{def:scale}, Remark~\ref{rem:scale_admissibility});
  \item Decomposition stability: no cancellation;
  \item Balance fixed-point: strict separation + $\asymp$-unavoidability;
  \item H1 exponent rigidity: injective map, log-inflation;
  \item Non-tautology, constant uniformity, method/operator scope;
  \item O1: Airy closed, general open.
\end{itemize}

\bigskip
\begin{quote}\itshape\large
The scale $c^{-\beta}(\log c)^{1+\gamma}$ is not merely optimal ---
it is the uniquely forced consequence of three mutually irreducible
constraints: exponent generation~(S1), $c$-covariance of the
asymptotic order~(S2), and diagonal projection~(S3).
Remove any one, and the balance fixed-point system
ceases to have a well-posed conclusion.
\end{quote}

\begin{thebibliography}{99}
\bibitem{paper18} \textit{Paper XVIII}, 2026.
\bibitem{paper19} \textit{Paper XIX}, 2026.
\bibitem{Kato1966} T.~Kato, \textit{Perturbation Theory for Linear Operators}, Springer, 1966.
\bibitem{Olver1997} F.~Olver, \textit{Asymptotics and Special Functions}, AK Peters, 1997.
\bibitem{Zettl2005} A.~Zettl, \textit{Sturm--Liouville Theory}, AMS, 2005.
\bibitem{ReedSimon2} M.~Reed, B.~Simon, \textit{Methods of Modern Mathematical Physics II},
  Academic Press, 1975.
\bibitem{Osipov2013} A.~Osipov, V.~Rokhlin, H.~Xiao,
  \textit{Prolate Spheroidal Wave Functions}, Springer, 2013.
\bibitem{BGT1987} N.~H.\ Bingham, C.~M.\ Goldie, J.~L.\ Teugels,
  \textit{Regular Variation}, Cambridge UP, 1987.
\bibitem{AbramowitzStegun} M.~Abramowitz, I.~A.\ Stegun,
  \textit{Handbook of Mathematical Functions}, NBS, 1964.
\end{thebibliography}

\end{document}
