\documentclass[12pt,a4paper]{amsart}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage[colorlinks=true,linkcolor=blue!60!black,citecolor=green!50!black,
            urlcolor=blue!70!black]{hyperref}
\usepackage[shortlabels]{enumitem}
\usepackage{booktabs}

%% ---- Theorem environments ----------------------------------------
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{problem}{Problem}[section]

%% ---- Notation ----------------------------------------------------
\newcommand{\PW}{\mathrm{PW}_\omega}
\newcommand{\Kop}{\mathcal{K}_c}
\newcommand{\GVM}{G^{(N)}_{\mathbf{p},c}}
\newcommand{\QM}{Q^{(N)}_{\mathbf{p},c}}
\newcommand{\psin}{\psi_n^{(c)}}
\newcommand{\EN}{E_{N,c}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\ip}[2]{\langle #1,\, #2\rangle}
\newcommand{\Rbulk}{\mathcal{R}_{\mathrm{bulk}}}
\newcommand{\Rturning}{\mathcal{R}_{\mathrm{turn}}}

\title[Frame Coercivity under First-Order Quadrature Control]
      {Frame Coercivity of Discrete Prolate Sampling Operators\\
       under First-Order Quadrature Control:\\
       Defect Decomposition and a Bulk--Turning Decay Mechanism}

\thanks{This version replaces the earlier conditional decay conjecture
by a proof architecture based on a quadrature--projection decomposition,
a fully quantified bulk phase-gap lemma, bulk-localised
integration-by-parts control, and a conservative turning-point estimate
via Cauchy--Schwarz and Airy-scale widths. The remaining quadrature
input is reduced to a standard first-order quadrature consistency
assumption, thereby replacing the earlier structural
quadrature-compatibility hypothesis by a verifiable error estimate.
A companion paper (Paper~II) develops scaling limits, trace formula,
and Weil-operator connections for the prolate framework;
see~\cite{PaperII}. A further companion (Paper~II\textsubscript{quad})
develops a separate PSWF-Gauss/XRY regime and formulates an implication
framework for oscillation-compatible quadrature based on PSWF-adapted
rules; see~\cite{PaperIIquad}. The semiclassical WKB amplitude and
spectral density results used in Sections~\ref{sec:decay} are
established unconditionally in Paper~IV~\cite{PaperIV}.}

\author{Anonymous Author}
\address{}
\email{}
\date{May 2026}

\subjclass[2020]{42C05, 42B10, 47B32, 41A60}

\keywords{prolate spheroidal wave functions, finite frames,
sampling in reproducing kernel Hilbert spaces,
Gram operator, Marcinkiewicz--Zygmund inequalities,
frame perturbation, banded decay, coercivity,
WKB asymptotics, turning points, quadrature error}

\begin{document}
\maketitle

\begin{abstract}
Let $\PW$ be the Paley--Wiener space of $\omega$-bandlimited functions
and $V_{N,c}:=\operatorname{span}\{\psi_0^{(c)},\dots,\psi_{N-1}^{(c)}\}$
the $N$-dimensional prolate subspace at time-bandwidth product $c=\omega T$.
We study the discrete sampling form
$\QM(f)=\sum_{j=1}^N w_j|f(p_j)|^2$ as a finite frame form on $V_{N,c}$.

We establish coercivity results for random sampling configurations and
for weighted sampling models satisfying a Marcinkiewicz--Zygmund lower
frame bound. Our main structural result is an exact defect decomposition
reducing frame analysis to simultaneous control of a weighted-frame
defect and a quadrature defect.

Assuming a standard first-order quadrature consistency estimate, we
derive a global quadrature defect bound in the PSWF eigenbasis:
\[
  |E_{mn}| \le C\,h c
  \qquad \text{for all } 0\le m,n\le N-1,
\]
where $h$ denotes the mesh size of the quadrature rule. The exact
identity $E_{mn} = R_{mn}^{\mathrm{quad}}$
(Proposition~\ref{prop:xry_split}) clarifies that the defect is purely
a quadrature error; the PSWF orthogonality term is exactly zero.
Separately, the continuous PSWF product integral exhibits
algebraic decay via a quantified phase-gap estimate and iterated
integration by parts, restricted to the bulk sector
$m,n\le\gamma_0 N_{\mathrm{Sh}}$ with $N_{\mathrm{Sh}}=2c/\pi$.

The bulk--turning analysis is developed as an independent structural
component; its role is to identify the oscillatory mechanism underlying
decay in continuous PSWF interactions and to motivate the interface
hypothesis required at the discrete level. The proof of the main
stability theorem does not depend on this analysis: the transfer of
continuous oscillatory structure to the discrete sampling level requires
a strictly additional assumption beyond first-order quadrature
consistency, encapsulated in the DSTP interface axiom and verified for
concrete quadrature schemes in a companion paper.
\end{abstract}

\tableofcontents

%% ==================================================================
\section{Introduction}
%% ==================================================================

The prolate spheroidal wave functions (PSWFs) form an orthonormal basis
for the finite-dimensional subspace $V_{N,c}$ of the Paley--Wiener
space $\PW$ that is optimally concentrated in a time--frequency window
\cite{Slepian1978,BonamiKaroui}. From the perspective of frame theory
and sampling in reproducing kernel Hilbert spaces, the weighted
evaluation map $f\mapsto(\sqrt{w_j}f(p_j))_j$ defines a finite frame
operator on $V_{N,c}$, whose associated Gram matrix encodes the
stability of the sampling scheme.

This paper addresses two questions. First, which classes of sampling
configurations yield coercive Gram operators on $V_{N,c}$? Second, how
does the discrete sampling Gram matrix compare to the reference spectral
matrix $G^{\mathrm{ref}}_{N,c}=\Kop|_{V_{N,c}}$ in the PSWF
eigenbasis?

For the first question we prove coercivity for random sampling and for
weighted MZ frames. For the second, we establish an exact algebraic
defect decomposition, derive a global first-order quadrature bound for
the discrete defect, and separately analyse the oscillatory decay
mechanism of the continuous PSWF product integral via a WKB--bulk /
Airy--turning decomposition restricted to the bulk sector. A companion
paper studies a PSWF-adapted XRY quadrature regime and conditional
spectral compatibility; see~\cite{PaperIIquad}.

The central contribution of this paper is therefore not a complete
frame stability theorem, but a \emph{reduction principle}: discrete
frame stability is reduced to a single verifiable spectral transfer
property (DSTP), and the precise obstruction---the failure of
continuous oscillatory decay to transfer to the discrete quadrature
level---is isolated and made explicit.

\subsection*{Main results}
The key new ingredient is the replacement of the earlier conditional
quadrature-compatibility hypothesis by a verifiable first-order
quadrature estimate. The proof architecture has four explicit
components:
\begin{enumerate}
\item \textbf{Quadrature--projection split}
  (Proposition~\ref{prop:xry_split}): $E_{mn}$ is a pure quadrature
  remainder; the PSWF orthogonality term is exactly zero.
\item \textbf{Spectral parameter control}
  (Proposition~\ref{prop:spectral_gap}): the transition-regime
  asymptotics for $\chi_n(c)$ are stated explicitly and used in the
  phase-gap argument.
\item \textbf{Uniform phase gap and IBP}
  (Lemmas~\ref{lem:bulk_spectral_sep}, \ref{lem:phase_gap_final},
  \ref{lem:bulk_derivative}, \ref{lem:ibp_uniform},
  and~\ref{lem:bulk_decay_continuous}): bulk-to-spectral separation
  makes the phase-gap estimate explicit in the bulk sector
  $m,n\le\gamma_0 N_{\mathrm{Sh}}$; bulk-restricted semiclassical
  derivative bounds control PSWF derivatives away from turning regions;
  bulk-localised iterated integration by parts yields algebraic decay
  for the continuous PSWF product integral.
\item \textbf{Quadrature reduction by derivative control}
  (Assumption~\ref{ass:quad_first_order},
  Lemma~\ref{lem:derivative_bound}, Theorem~\ref{thm:banded_decay}):
  the quadrature remainder is controlled by a standard first-order
  error term involving $\int|f'_{mn}|$, estimated via PSWF regularity.
\end{enumerate}

The oscillatory bulk--turning analysis in component~3 is developed as
an independent structural unit. It is \emph{not} used in the proof of
the main stability theorem (Theorem~\ref{thm:conditional_frame_schur});
its role is to prove that continuous PSWF product decay is a
\emph{purely continuous phenomenon}: it identifies the analytic decay
mechanism of PSWF interactions and demonstrates that this structure is
destroyed by generic quadrature. Section~\ref{sec:decay} therefore
constitutes a negative result in this sense, motivating DSTP as the
necessary interface axiom. That gap is formalised by the Discrete
Spectral Transfer Property (DSTP), which serves as the explicit
axiomatic interface between the present paper and~\cite{PaperIIquad}.

\subsection*{Organisation}
Section~\ref{sec:setup} fixes notation.
Section~\ref{sec:coercivity} proves coercivity.
Section~\ref{sec:decay} establishes the decay mechanism and its
separation from the discrete level.
Section~\ref{sec:stability} derives the combined frame stability theorem.
Section~\ref{sec:bridge} explains the conceptual boundary between the
present paper and the companion framework.
Section~\ref{sec:open} lists open problems.
Appendix~\ref{app:airy} contains the Airy approximation lemma.

%% ==================================================================
\section{Setup and notation}\label{sec:setup}
%% ==================================================================

\subsection{Paley--Wiener space and concentration operator}

We use the Fourier convention
$\widehat{f}(\xi)=\int_{\mathbb{R}}f(x)e^{-2\pi ix\xi}\,dx$.
The Paley--Wiener space $\PW$ consists of all $f\in L^2(\mathbb{R})$
with Fourier support in $[-\omega/(2\pi),\omega/(2\pi)]$. It is a
reproducing-kernel Hilbert space with kernel
\[
  K_{\PW}(x,y)
  =\tfrac{\omega}{\pi}\operatorname{sinc}(\omega(x-y)/\pi),
  \qquad K_{\PW}(x,x)=\omega/\pi
\]
\cite{BonamiKaroui}.

The prolate concentration operator is
\[
  \Kop f(x)=\int_{-T}^{T}\frac{\sin\omega(x-y)}{\pi(x-y)}f(y)\,dy,
  \qquad \Kop:\PW\to\PW,
\]
with eigenvalues $1>\lambda_0(c)>\lambda_1(c)>\cdots>0$, PSWF
eigenfunctions $\{\psin\}_{n\ge 0}$ forming an orthonormal basis of
$\PW$ \cite{Slepian1978}, and associated spectral parameters $\chi_n(c)$
(eigenvalues of the PSWF differential operator; see
\cite{OsipovRokhlinXiao,BonamiKaroui2014}). We set $c=\omega T$ and
$V_{N,c}:=\operatorname{span}\{\psi_0^{(c)},\dots,\psi_{N-1}^{(c)}\}$.
The transition regime $N\sim c$ is the natural operating range.
We denote the Shannon number by $N_{\mathrm{Sh}}:=2c/\pi$ and fix
$\gamma_0\in(0,1)$; the \emph{bulk sector} is $\{0\le k\le\gamma_0 N_{\mathrm{Sh}}\}$.

\subsection{Frame operator and defect matrix}

\begin{definition}[Weighted evaluation frame and Gram matrix]%
\label{def:GramForm}
For distinct $\mathbf{p}=(p_1,\dots,p_N)\subset[-T,T]$ and weights
$w_j>0$, the evaluation matrix is
$M(\mathbf{p},c)_{jn}:=\psi_n^{(c)}(p_j)$ and
$D:=\operatorname{diag}(w_1,\dots,w_N)$. The weighted Gram matrix is
\[
  G^{\mathrm{w}}_{N,c}(\mathbf{p},D)
  :=M(\mathbf{p},c)^*D\,M(\mathbf{p},c).
\]
For $f=\sum_{n=0}^{N-1}a_n\psin\in V_{N,c}$,
$\QM(f):=\sum_{j=1}^N w_j|f(p_j)|^2
=\ip{G^{\mathrm{w}}_{N,c}\mathbf{a}}{\mathbf{a}}$.
\end{definition}

\begin{definition}[Reference spectral matrix and defect]%
\label{def:ref_matrix}
\[
  G^{\mathrm{ref}}_{N,c}
  :=\operatorname{diag}(\lambda_0(c),\dots,\lambda_{N-1}(c)),
  \qquad
  \EN:=G^{\mathrm{w}}_{N,c}(\mathbf{p},D)-G^{\mathrm{ref}}_{N,c}.
\]
The defect entries are
\[
  E_{mn}
  :=\sum_{j=1}^N w_j\,\psi_m^{(c)}(p_j)\overline{\psi_n^{(c)}(p_j)}
    -\lambda_n(c)\,\delta_{mn}.
\]
\end{definition}

\begin{definition}[Frame-stable regime]\label{def:framestable}
$\mathbf{p}$ is in the frame-stable regime if
$\norm{\EN}_2<\lambda_{N-1}(c)$.
\end{definition}

\begin{proposition}[Spectral parameter asymptotics]%
\label{prop:spectral_gap}
For $n$ in the bulk sector $n \le \gamma_0 N_{\mathrm{Sh}}$, the
PSWF spectral parameters satisfy
\[
  \chi_n(c) = n(n+1) + \frac{(2n+1)c}{\pi} + O\!\left(\frac{c^2}{n}\right),
\]
as established in \cite[Section~2.5]{OsipovRokhlinXiao}.
In particular, the leading-order spacing
\[
  \chi_{n+1}(c) - \chi_n(c) \sim \frac{2c}{\pi} + O(n)
\]
is dominated by the linear-in-$c$ term for $n \ll c$,
and is dominated by the linear-in-$n$ term for $n \gg 1$.
The operative bound used in this paper is the uniform monotonicity
$\chi_n(c)$ strictly increasing in $n$ \cite[Section~2.5]{OsipovRokhlinXiao},
combined with the Prüfer-phase lower bound
$\omega_n^{\mathrm{cl}}(x) \ge c_{\delta,\gamma}\,n$
on the bulk interval (Lemma~\ref{lem:phase_gap_final}, Step~1),
which together imply a uniform spectral gap of order $|m-n|$
in the bulk sector. No global linear-in-$c$ spacing is
claimed outside this regime.
\end{proposition}

%% ==================================================================
\section{Unisolvence and coercivity}\label{sec:coercivity}
%% ==================================================================

\begin{definition}[Frame condition]\label{def:frame}
$\{(p_j,w_j)\}_{j=1}^N$ is a weighted frame for $V_{N,c}$ if
$A\norm{f}_{L^2}^2\le\sum_j w_j|f(p_j)|^2\le B\norm{f}_{L^2}^2$
for all $f\in V_{N,c}$ with $0<A\le B<\infty$.
\end{definition}

\begin{definition}[Unisolvent set]
$\mathbf p$ is unisolvent for $V_{N,c}$ if
$f\mapsto(f(p_1),\dots,f(p_N))$ is injective on $V_{N,c}$.
\end{definition}

\begin{lemma}[Identity principle]\label{lem:identity}
If $f\in\PW$ vanishes on a set with a limit point in $\mathbb R$,
then $f\equiv 0$.
\end{lemma}
\begin{proof}
Every $f\in\PW$ is entire of exponential type; apply the identity
theorem.
\end{proof}

\begin{proposition}[MZ lower bound implies unisolvence]%
\label{prop:mz_unisolvent}
If $\{(p_j,w_j)\}$ is a weighted frame with lower bound $A>0$, then
$\mathbf p$ is unisolvent for $V_{N,c}$.
\end{proposition}
\begin{proof}
$f(p_j)=0$ for all $j$ implies $A\norm{f}^2\le 0$, hence $f\equiv 0$.
\end{proof}

\begin{proposition}[Random sampling]\label{prop:random}
Let $p_1,\dots,p_N$ be i.i.d.\ absolutely continuous on $[-T,T]$.
Then $\mathbf{p}$ is unisolvent for $V_{N,c}$ almost surely.
\end{proposition}
\begin{proof}
$(p_1,\dots,p_N)\mapsto\det M(\mathbf{p},c)$ is real-analytic and
non-identically zero, so its zero set has measure zero.
\end{proof}

\begin{lemma}[Pointwise RKHS bound]\label{lem:upper}
For $f\in\PW$ and $x\in\mathbb R$,
$|f(x)|^2\le(\omega/\pi)\norm{f}_{L^2}^2$.
\end{lemma}
\begin{proof}
Cauchy--Schwarz with the reproducing kernel.
\end{proof}

\begin{theorem}[Frame coercivity]\label{thm:coercivity}
Let $\mathbf{p}$ be unisolvent for $V_{N,c}$ and $D=I$. Then
\[
  \sqrt{\lambda_{\min}(\GVM)}\,\norm{f}_{L^2}
  \le\QM(f)^{1/2}
  \le\sqrt{N\omega/\pi}\,\norm{f}_{L^2}
  \qquad\forall f\in V_{N,c}.
\]
\end{theorem}
\begin{proof}
Lower bound: Rayleigh quotient. Upper bound: sum Lemma~\ref{lem:upper}
over nodes.
\end{proof}

\begin{lemma}[Operator norm bound]\label{lem:lambda_max_bound}
$\lambda_{\max}(G^{\mathrm{w}}_{N,c})\le N\omega/\pi$.
\end{lemma}

%% ==================================================================
\section{Quadrature decomposition and oscillatory decay}\label{sec:decay}
%% ==================================================================

\begin{proposition}[Quadrature--Projection Decomposition]%
\label{prop:xry_split}
Let $M(\mathbf{p},c)$ be the evaluation matrix and
$D=\mathrm{diag}(w_j)$. Then
\[
  E_{mn}
  =
  \underbrace{
  \Bigl(
  \int_{-T}^{T}\psi_m^{(c)}(x)\overline{\psi_n^{(c)}(x)}\,dx
  -\lambda_n(c)\delta_{mn}
  \Bigr)}_{=\,0}
  +
  R_{mn}^{\mathrm{quad}},
\]
where the quadrature remainder is
\[
  R_{mn}^{\mathrm{quad}}
  :=
  \sum_{j=1}^N w_j f_{mn}(p_j)
  -\int_{-T}^{T}f_{mn}(x)\,dx,
  \qquad f_{mn}(x)
  :=\psi_m^{(c)}(x)\overline{\psi_n^{(c)}(x)}.
\]
Since $\Kop$ is self-adjoint on $L^2([-T,T])$ and $\{\psi_n^{(c)}\}$
is an orthonormal eigenbasis of the concentration operator restricted
to $L^2([-T,T])$, the eigenvalue identity
$\Kop\psi_n^{(c)}=\lambda_n(c)\psi_n^{(c)}$ together with the kernel
representation of $\Kop$ yields directly
\[
  \int_{-T}^{T}\psi_n^{(c)}(x)\overline{\psi_m^{(c)}(x)}\,dx
  = \lambda_n(c)\,\delta_{mn},
\]
which proves that the bracketed term vanishes. Consequently,
\[
  E_{mn} = R_{mn}^{\mathrm{quad}},
\]
and all later estimates use $|E_{mn}| = |R_{mn}^{\mathrm{quad}}|$.
\end{proposition}

\begin{remark}
Proposition~\ref{prop:xry_split} makes explicit that $E_{mn}$ is
\emph{purely} a quadrature error for the PSWF product integral. This
isolates the single error carrier in the frame defect and shows that
continuous PSWF orthogonality contributes nothing to $E_{mn}$.
\end{remark}

\begin{definition}[Bulk and turning regions]\label{def:bulk}
Let $x_n^*>0$ denote the positive turning point of $\psi_n^{(c)}$,
defined by $c^2(x_n^*)^2/T^2 = \chi_n(c)$,
and fix $\rho>0$. The Airy neighbourhood is
\[
  U_n
  :=\{x\in[-T,T]:|x-x_n^*|\le\rho c^{-1/3}\}
   \cup\{x\in[-T,T]:|x+x_n^*|\le\rho c^{-1/3}\}.
\]
For a pair $(m,n)$:
\[
  \Rturning^{(m,n)}:=U_m\cup U_n,
  \qquad
  \Rbulk^{(m,n)}:=[-T,T]\setminus\Rturning^{(m,n)}.
\]
\end{definition}

\begin{remark}
The width $O(c^{-1/3})$ is the standard Airy transition scale
\cite{Olver1974}: beyond the neighbourhoods of the turning points
$\pm x_n^*$ the WKB phase is non-stationary for $|m-n|\ge 1$.
The turning point $x_n^*$ is well-defined for $n\le\gamma_0 N_{\mathrm{Sh}}$
by the strict inequality $\chi_n(c) < c^2$ in the bulk sector
\cite[Section~2.3]{OsipovRokhlinXiao}.
\end{remark}

\begin{lemma}[Bulk spectral separation via turning-point geometry]%
\label{lem:bulk_spectral_sep}
Let $x\in\Rbulk^{(m,n)}$. Then there exists $\delta_0>0$, depending
only on $\rho$, such that
\[
  \left|\chi_m(c)-\frac{c^2x^2}{T^2}\right|\ge\delta_0 c^{2/3},
  \qquad
  \left|\chi_n(c)-\frac{c^2x^2}{T^2}\right|\ge\delta_0 c^{2/3}.
\]
\end{lemma}
\begin{proof}
For $k\in\{m,n\}$, the turning point $x_k^*$ satisfies
$\chi_k(c)=c^2(x_k^*)^2/T^2$ by definition (Definition~\ref{def:bulk}),
so
$\chi_k(c)-c^2x^2/T^2=(c^2/T^2)((x_k^*)^2-x^2)=(c^2/T^2)(x_k^*-x)(x_k^*+x)$.
On $\Rbulk^{(m,n)}$, Definition~\ref{def:bulk} ensures
$|x_k^*\pm x|\ge\rho c^{-1/3}$, hence
\[
  \left|\chi_k(c)-\frac{c^2x^2}{T^2}\right|
  \ge\frac{c^2}{T^2}\,\rho^2 c^{-2/3}
  =\frac{\rho^2}{T^2}c^{2/3}.
\]
Set $\delta_0:=\rho^2/T^2$.
\end{proof}

\begin{lemma}[Uniform phase gap in the bulk sector]%
\label{lem:phase_gap_final}
Let $m,n$ lie in the bulk sector $m,n\le\gamma_0 N_{\mathrm{Sh}}$,
and fix $\rho>0$. There exist $c_0>0$ and $\kappa>0$
such that for all $c\ge c_0$, uniformly in
$0\le m\ne n\le\gamma_0 N_{\mathrm{Sh}}$:
\[
  |\Phi_{mn}'(x)|
  \;\ge\;
  \frac{\kappa}{cT}\,|m-n|
  \qquad\forall x\in\Rbulk^{(m,n)}.
\]
The lemma applies only in the bulk sector; the transition regime
$|k-N_{\mathrm{Sh}}|=O(c^{1/3})$ is excluded.
\end{lemma}
\begin{proof}
Set $\mu:=\chi_m(c)-c^2x^2/T^2$, $\nu:=\chi_n(c)-c^2x^2/T^2$.
By Lemma~\ref{lem:bulk_spectral_sep}, $\mu,\nu\ge\delta_0 c^{2/3}$.
Using $|\sqrt\mu-\sqrt\nu|=|\mu-\nu|/(\sqrt\mu+\sqrt\nu)$:
\[
  |\Phi_{mn}'(x)|
  =\frac{|\chi_m(c)-\chi_n(c)|}{T(\sqrt\mu+\sqrt\nu)}.
\]

\noindent\textbf{Step~1 (numerator): spectral monotonicity and
Prüfer-phase lower bound.}
By \cite[Section~2.5]{OsipovRokhlinXiao}, $\chi_n(c)$ is strictly
increasing in $n$ for fixed $c$. To obtain a uniform lower bound on
the gap, we use the Prüfer-phase representation from
\cite[Lemma~3.1]{PaperIV}: on the bulk interval $I_\delta$, the
classical frequency satisfies
\[
  \omega_n^{\mathrm{cl}}(x)
  \;:=\;\sqrt{\chi_n(c)/T^2 - c^2x^2/T^2 + c^2}
  \;\ge\; c_{\delta,\gamma}\,n
\]
uniformly for $n \le \gamma_0 N_{\mathrm{Sh}}$.
By the Bohr--Sommerfeld identity \cite[Lemma~3.2]{PaperIV},
\[
  \int_{-x_+(n)}^{x_+(n)} \omega_n^{\mathrm{cl}}(x)\,dx
  \;=\; \pi\!\left(n+\tfrac{1}{2}\right) + O\!\left(\tfrac{1}{n}\right),
\]
so consecutive turning-point integrals differ by $\pi + O(1/n)$.
Combining with the mean value theorem applied to
$n \mapsto \chi_n(c)$ (which is differentiable and monotone increasing),
there exists $\gamma' > 0$ depending only on $\delta$, $\gamma_0$ such that
\[
  \chi_{n+1}(c) - \chi_n(c) \;\ge\; \gamma'
  \qquad\text{for all } n \le \gamma_0 N_{\mathrm{Sh}},\; c \ge c_0.
\]
Summing over consecutive indices:
\[
  |\chi_m(c) - \chi_n(c)| \;\ge\; \gamma'\,|m-n|.
\]

\noindent\textbf{Step~2 (denominator).}
$\sqrt\mu+\sqrt\nu \le 2\sqrt{\chi_{N-1}(c)} \le C_3\,c$
since $\chi_{N-1}(c)\lesssim c^2$ in the regime $N\sim c$.

\noindent\textbf{Step~3 (assembly).}
Combining Steps~1 and~2:
\[
  |\Phi_{mn}'(x)|
  \;\ge\; \frac{\gamma'\,|m-n|}{T \cdot C_3\,c}
  \;=:\; \frac{\kappa\,|m-n|}{cT}.
\]

\noindent\textbf{Remark on IBP scaling.}
The phase gap scales as $|m-n|/(cT)$, which is weaker by one factor of
$c$ compared to a hypothetical linear-in-$c$ gap. This is consistent
with the quadrature defect scaling $|E_{mn}| \le Chc$: the additional
factor $c$ in the denominator of the phase gap is absorbed by the $hc$
prefactor in DSTP (Definition~\ref{def:dstp}), and does not affect the
Schur summability of $(1+|m-n|)^{-p}$ used in
Theorem~\ref{thm:conditional_frame_schur}.
\end{proof}

\begin{remark}[Bulk restriction and transition regime]
\label{rem:phase_gap_regime}
Lemma~\ref{lem:phase_gap_final} is restricted to the bulk sector
$m,n\le\gamma_0 N_{\mathrm{Sh}}$. In the transition regime
$|k-N_{\mathrm{Sh}}|=O(c^{1/3})$, the spectral gap degrades to
$O(c^{1/3})$ and the argument above does not apply; that regime
requires Airy-scale analysis and is excluded throughout
Section~\ref{sec:decay}.
\end{remark}

\begin{lemma}[Bulk semiclassical derivative control]%
\label{lem:bulk_derivative}
Fix $\rho>0$ and $\gamma_0\in(0,1)$. For every integer $k\ge 0$
there exists $C_k>0$, depending only on $k$, $\rho$, and $\gamma_0$,
such that
\[
  \|\partial_x^k\psi_n^{(c)}\|_{L^\infty(\Rbulk^{(m,n)})}
  \le C_k\,c^{k/2}
\]
uniformly for $0\le n\le \gamma_0 N_{\mathrm{Sh}}$ and $c\ge 1$.
\end{lemma}
\begin{proof}
On $\Rbulk^{(m,n)}$ the PSWF admits a WKB decomposition
$\psi_n^{(c)}(x)=r_n(x)\sin(\theta_n(x))+O(c^{-1})$,
where $\theta_n'(x)\sim c/T$ is the local phase derivative
and $r_n(x)$ is the WKB amplitude \cite[Ch.~11]{Olver1974}.
By \cite[Lemmas~5.3,~5.4]{PaperIV}, the amplitude satisfies
$r_n(x)^2=2\lambda_n(c)\rho^{\mathrm{cl}}(x)(1+O(1/n))$
uniformly for $n\le\gamma_0 N_{\mathrm{Sh}}$,
with $\rho^{\mathrm{cl}}(x)\le c/(\pi T)$ in the bulk sector.
The $c^{k/2}$ scaling reflects the WKB balance between phase and
amplitude contributions: each differentiation of $\sin(\theta_n)$
produces a phase factor $\theta_n'\sim c/T$, while the amplitude
normalization $r_n\sim (c/T)^{-1/2}$ (from
$r_n^2\lesssim \lambda_n c/T\le c/T$ and
$\rho^{\mathrm{cl}}\le c/(\pi T)$) introduces a compensating
$c^{-1/2}$, yielding a net $c^{1/2}$ per order of differentiation
in the leading semiclassical scaling in the bulk regime,
uniformly away from turning points by $\rho c^{-1/3}$.
Derivatives of the amplitude $r_n$ contribute lower-order terms in $c$
and do not affect the leading semiclassical scaling in the bulk regime.
By \cite{BonamiKaroui2014}, uniform $C^0$ bounds hold for $n\le N-1$
on compact subsets of the bulk separated from turning points by
Airy-scale neighbourhoods of size $\rho c^{-1/3}$; combined with the
above, this gives uniformity across $n\le\gamma_0 N_{\mathrm{Sh}}$.
\end{proof}

\begin{lemma}[Bulk-localised uniform IBP control]\label{lem:ibp_uniform}
Let $a_{mn}(x)=A_m(x)\overline{A_n(x)}$ be the WKB amplitude product
and $L:=(i\Phi_{mn}'(x))^{-1}\partial_x$. For every $p\ge 1$,
\[
  \|(L^*)^p a_{mn}\|_{L^1(\Rbulk^{(m,n)})}
  \le C_p\left(\frac{c}{|m-n|}\right)^p.
\]
Moreover, $(L^*)^p a_{mn}$ is a finite linear combination of terms
$c_{p,k,\ell}(x)\,\partial_x^k a_{mn}(x)/(\Phi_{mn}'(x))^{\ell}$,
$0\le k\le p$, $1\le\ell\le p$, with coefficients depending only on
$p$ and finitely many derivatives of $\Phi_{mn}'$.
The bound holds uniformly for all $p\le p_0$ (fixed finite order).

At the boundary $\partial\Rbulk^{(m,n)}$, the WKB and Airy
representations of $\psi_n^{(c)}$ are matched via the standard
uniform asymptotic connection formula \cite[Ch.~11]{Olver1974},
which provides a uniform asymptotic matching between WKB and Airy
regimes, ensuring continuity of amplitude estimates across
$\partial\Rbulk^{(m,n)}$; boundary contributions are therefore
interpreted as point evaluations at the Airy interface, legitimised
by this matching.
\end{lemma}
\begin{proof}
The formal adjoint is
$L^*=-(i\Phi_{mn}')^{-1}\partial_x+b_1(x)(\Phi_{mn}')^{-2}$,
with $b_1$ smooth on $\Rbulk^{(m,n)}$. By induction, $(L^*)^p a_{mn}$
has the stated structure. By Lemma~\ref{lem:phase_gap_final},
$|(\Phi_{mn}')^{-1}|\le cT/(\kappa|m-n|)$.
Lemma~\ref{lem:bulk_derivative} gives uniform $C^k$ bounds for
amplitude derivatives on $\Rbulk^{(m,n)}$.

Each integration-by-parts step produces a boundary term at
$\partial\Rbulk^{(m,n)}$, i.e., at spatial distance
$\rho c^{-1/3}$ from the turning points. These terms are
point evaluations at the Airy interface; the WKB--Airy matching
stated above ensures they are bounded by
\[
  \bigl|a_{mn}(x)\cdot(\Phi_{mn}'(x))^{-1}\bigr|_{x\in\partial\Rbulk^{(m,n)}}
  \;\le\;
  C_\infty^2\cdot\frac{cT}{\kappa\,|m-n|},
\]
using $|a_{mn}(x)|\le C_\infty^2$ uniformly on $\partial\Rbulk^{(m,n)}$
by \cite{BonamiKaroui2014}.
After $p$ steps, the accumulated boundary contribution is bounded
uniformly in $p\le p_0$ by
$C_p'\,c^{-1/3}\cdot(c/|m-n|)^p$,
where the factor $c^{-1/3}$ arises from the measure
$|\partial\Rbulk^{(m,n)}|=O(c^{-1/3})$ of the excluded Airy neighbourhood.
This is absorbed into the turning-region estimate $O(c^{-1/3})$
of Lemma~\ref{lem:turning_revised}, uniformly for all $p\le p_0$.

Integrating over $\Rbulk^{(m,n)}\subset[-T,T]$ yields the stated
$L^1$ bound.
\end{proof}

\begin{lemma}[Bulk decay for the continuous integral]\label{lem:bulk_decay}
Under the hypotheses of Lemma~\ref{lem:ibp_uniform}, for $m\neq n$
with $m,n\le\gamma_0 N_{\mathrm{Sh}}$,
\[
  \left|\int_{\Rbulk^{(m,n)}}f_{mn}(x)\,dx\right|
  \le C_p\left(\frac{c}{|m-n|}\right)^p
  \le C_p'(1+|m-n|)^{-p},
\]
with constants independent of $m,n,N,c$.
\end{lemma}
\begin{proof}
Apply $L$ iteratively on $\Rbulk^{(m,n)}$. Boundary terms are
controlled via Lemma~\ref{lem:ibp_uniform}.
\end{proof}

\begin{lemma}[Turning-point estimate]\label{lem:turning_revised}
For all $0\le m,n\le N-1$,
$\int_{U_m}|\psi_m^{(c)}(x)\psi_n^{(c)}(x)|\,dx\le C\,c^{-1/3}$.
\end{lemma}
\begin{proof}
By \cite{BonamiKaroui2014}, $\|\psi_k^{(c)}\|_{L^\infty}\le C_\infty$
uniformly. Since $|U_m|=O(c^{-1/3})$,
Cauchy--Schwarz gives
$\int_{U_m}|\psi_m^{(c)}\psi_n^{(c)}|\le C_\infty^2|U_m|=O(c^{-1/3})$.
\end{proof}

\begin{assumption}[First-order quadrature consistency]%
\label{ass:quad_first_order}
Let $\{(p_j,w_j)\}_{j=1}^N$ be a quadrature rule on $[-T,T]$ with
mesh size $h:=\max_{j}|p_{j+1}-p_j|$. We assume there exists $C_Q>0$
such that for all $f\in W^{1,1}([-T,T])$,
\[
  \Bigl|\int_{-T}^{T}f(x)\,dx-\sum_{j=1}^N w_jf(p_j)\Bigr|
  \le C_Q\,h\int_{-T}^{T}|f'(x)|\,dx.
\]
\end{assumption}

\begin{remark}[Scope of the quadrature input]\label{rem:scope_quad}
Assumption~\ref{ass:quad_first_order} is intentionally coarse: it
controls only the absolute size of the quadrature remainder and encodes
no oscillation-sensitive cancellation, PSWF-adapted exactness, or
transfer of continuous off-diagonal decay to the discrete level.
\end{remark}

\begin{lemma}[Derivative bound for PSWF products]%
\label{lem:derivative_bound}
For $f_{mn}(x):=\psi_m^{(c)}(x)\overline{\psi_n^{(c)}(x)}$, if the
Bonami--Karoui bounds $\|\psi_k^{(c)}\|_{L^\infty}\le C_0$ and
$\|\psi_k^{(c)\,\prime}\|_{L^\infty}\le C_1 c$ hold uniformly for
$0\le k\le N-1$ \cite{BonamiKaroui2014}, then
\[
  \int_{-T}^{T}|f'_{mn}(x)|\,dx\le Cc,
\]
with a constant independent of $m,n,N,c$.
\end{lemma}
\begin{proof}
Differentiate: $|f'_{mn}|\le|\psi_m'||\psi_n|+|\psi_m||\psi_n'|
\le 2C_0C_1 c$. Integration over $[-T,T]$ gives the claim.
\end{proof}

\begin{theorem}[Quadrature defect bound]\label{thm:banded_decay}
Let Assumption~\ref{ass:quad_first_order} hold. Then
$|E_{mn}|\le Chc$ for all $0\le m,n\le N-1$,
with $C>0$ independent of $m,n,N,c$.
\end{theorem}
\begin{proof}
By Proposition~\ref{prop:xry_split}, $|E_{mn}|=|R_{mn}^{\mathrm{quad}}|$.
Apply Assumption~\ref{ass:quad_first_order} and then
Lemma~\ref{lem:derivative_bound}:
$|E_{mn}|\le C_Q h\int|f'_{mn}|\le C_QCh c$.
\end{proof}

\begin{remark}[Sharpness]\label{rem:sharpness_quad}
The bound $|E_{mn}|\le Chc$ is structure-blind. Without additional
oscillation-sensitive quadrature structure, no off-diagonal decay in
$|m-n|$ is expected at the discrete level; the estimate is essentially
sharp under the present assumptions.
\end{remark}

\begin{lemma}[Continuous bulk--turning decay]%
\label{lem:bulk_decay_continuous}
For all $p\ge 1$ and $0\le m,n\le\gamma_0 N_{\mathrm{Sh}}$ with $m\ne n$,
\[
  \left|\int_{-T}^{T}\psi_m^{(c)}(x)\overline{\psi_n^{(c)}(x)}\,dx\right|
  \le C_p(1+|m-n|)^{-p}+C\,c^{-1/3}.
\]
For $m=n$ the integral equals $\lambda_n(c)$.
\end{lemma}
\begin{proof}
For $m\neq n$ decompose over $\Rbulk^{(m,n)}$ and $\Rturning^{(m,n)}$.
Lemma~\ref{lem:bulk_decay} bounds the bulk term by
$C_p(1+|m-n|)^{-p}$; Lemma~\ref{lem:turning_revised} bounds the
turning term by $Cc^{-1/3}$.
\end{proof}

\begin{remark}
This decay describes the continuous PSWF kernel in the bulk sector
and is a purely continuous phenomenon. It does \emph{not} transfer to
$E_{mn}$ without additional oscillation-compatible quadrature
assumptions: generic quadrature rules destroy this structure, yielding
only the structure-blind bound $|E_{mn}| \le Chc$. The gap between
continuous decay and discrete frame stability is precisely what DSTP
(Definition~\ref{def:dstp}) is designed to close;
see~\cite{PaperIIquad}.
\end{remark}

%% ==================================================================
\section{Defect decomposition and quantitative frame stability}%
\label{sec:stability}
%% ==================================================================

\begin{theorem}[Defect decomposition]\label{thm:main_reduction}
For any $(\mathbf p,D)$ with $D>0$,
\[
  \EN=M(\mathbf p,c)^*(D-I)M(\mathbf p,c)
     +(G^{\mathrm{w}}_{N,c}(\mathbf{p},I)-G^{\mathrm{ref}}_{N,c}),
\]
and $\norm{\EN}_2\le\norm{D-I}_2\cdot N\omega/\pi
+\norm{G^{\mathrm{w}}_{N,c}(\mathbf{p},I)-G^{\mathrm{ref}}_{N,c}}_2$.
\end{theorem}
\begin{proof}
By Definition~\ref{def:ref_matrix},
\[
  \EN = G^{\mathrm{w}}_{N,c}(\mathbf{p},D) - G^{\mathrm{ref}}_{N,c}
      = M(\mathbf{p},c)^* D\, M(\mathbf{p},c) - G^{\mathrm{ref}}_{N,c}.
\]
Add and subtract $M(\mathbf{p},c)^* M(\mathbf{p},c)$:
\[
  \EN = M(\mathbf{p},c)^*(D - I)M(\mathbf{p},c)
      + \bigl(M(\mathbf{p},c)^* M(\mathbf{p},c) - G^{\mathrm{ref}}_{N,c}\bigr).
\]
Since $M(\mathbf{p},c)^* M(\mathbf{p},c) = G^{\mathrm{w}}_{N,c}(\mathbf{p},I)$,
the second bracket equals
$G^{\mathrm{w}}_{N,c}(\mathbf{p},I) - G^{\mathrm{ref}}_{N,c}$,
yielding the claimed decomposition.
For the operator norm bound,
\begin{align*}
  \norm{\EN}_2
  &\le \norm{M(\mathbf{p},c)^*(D-I)M(\mathbf{p},c)}_2
     + \norm{G^{\mathrm{w}}_{N,c}(\mathbf{p},I) - G^{\mathrm{ref}}_{N,c}}_2 \\
  &\le \norm{D - I}_2 \cdot \norm{M(\mathbf{p},c)}_2^2
     + \norm{G^{\mathrm{w}}_{N,c}(\mathbf{p},I) - G^{\mathrm{ref}}_{N,c}}_2.
\end{align*}
By Lemma~\ref{lem:lambda_max_bound},
$\norm{M(\mathbf{p},c)}_2^2 = \lambda_{\max}(M^*M) \le N\omega/\pi$,
which completes the proof.
\end{proof}

\begin{lemma}[Weyl perturbation]\label{lem:weyl}
If $\norm{\EN}_2<\lambda_{N-1}(c)$ then
$\lambda_{\min}(G^{\mathrm{w}}_{N,c})
\ge\lambda_{N-1}(c)-\norm{\EN}_2>0$.
\end{lemma}
\begin{proof}
$\lambda_{\min}(A+E)\ge\lambda_{\min}(A)-\norm{E}_2$ for Hermitian
matrices.
\end{proof}

\begin{definition}[Discrete Spectral Transfer Property]\label{def:dstp}
Let $\{(p_j,w_j)\}_{j=1}^N$ satisfy Assumption~\ref{ass:quad_first_order}
with defect matrix $\EN$. We say the rule satisfies the \emph{Discrete
Spectral Transfer Property} (DSTP) on $V_{N,c}$ if there exist
$C>0$ and $p>1$ such that
\[
  |E_{mn}|\le Chc\,(1+|m-n|)^{-p}
  \qquad\text{for all }0\le m,n\le N-1.
\]
DSTP is an \emph{axiomatic interface hypothesis} that goes strictly
beyond Assumption~\ref{ass:quad_first_order}: it asserts that the
discrete defect inherits a Schur-summable off-diagonal profile at
scale $hc$, in a form that mirrors the continuous oscillatory decay
established in Section~\ref{sec:decay}. DSTP is \emph{not} a
consequence of first-order quadrature consistency, nor is it implied
by the continuous decay estimates of Section~\ref{sec:decay}. It is
an \emph{additional structural assumption} that precisely encodes the
missing transfer mechanism between continuous oscillatory decay and
discrete frame stability, and whose verification for concrete
quadrature schemes is deferred to~\cite{PaperIIquad}.
\end{definition}

\begin{remark}[DSTP: context, scope, and heuristics]%
\label{rem:dstp_literature}
DSTP is not expected to hold for classical quadrature families such as
Riemann sums, Newton--Cotes rules, or standard Gauss--Legendre rules
applied without adaptation to the PSWF eigenbasis: these rules carry
no oscillation-alignment with respect to the PSWF spectral structure
and therefore yield only the structure-blind bound of
Theorem~\ref{thm:banded_decay}. DSTP should instead be viewed as a
structural strengthening of Marcinkiewicz--Zygmund-type quadrature
conditions in the spectral regime, analogous in spirit to classical
exactness and stability requirements in MZ theory
\cite{OsipovRokhlinXiao,Xiao2001}, but specialised to the
off-diagonal decay geometry of PSWF kernels.

Heuristically, DSTP can be interpreted as a discrete analogue of
stationary-phase cancellation: it holds precisely when the sampling
rule is \emph{phase-aware} with respect to the PSWF eigenbasis, in
the sense that the quadrature weights align with the oscillatory
structure of $f_{mn}$. The existence and explicit construction of
rules satisfying DSTP is the subject of~\cite{PaperIIquad}.
\end{remark}

\begin{theorem}[Conditional frame stability under Schur-decay hypothesis]%
\label{thm:conditional_frame_schur}
Let $\{(p_j,w_j)\}_{j=1}^N$ satisfy Assumption~\ref{ass:quad_first_order}
with $N\sim c$, and assume DSTP (Definition~\ref{def:dstp}). If
\[
  Chc\sum_{k\in\mathbb Z}(1+|k|)^{-p}<\lambda_{N-1}(c),
\]
then $\{(p_j,w_j)\}$ is a weighted frame for $V_{N,c}$ with bounds
\begin{align*}
  A(N,c)&:=\lambda_{N-1}(c)-Chc\!\sum_{k}(1+|k|)^{-p}>0,\\
  B(N,c)&:=\lambda_0(c)+Chc\!\sum_{k}(1+|k|)^{-p}.
\end{align*}
\end{theorem}
\begin{proof}
DSTP and the Schur test give
$\norm{\EN}_2\le Chc\sum_k(1+|k|)^{-p}$.
Apply Lemma~\ref{lem:weyl} for the lower bound and the reference
diagonal for the upper bound.
\end{proof}

\begin{corollary}\label{cor:fixed_param_scope}
Theorem~\ref{thm:conditional_frame_schur} yields finite-dimensional
frame stability under DSTP and first-order quadrature control.
Assumption~\ref{ass:quad_first_order} alone does not imply the
Schur-summable structure required.
\end{corollary}

\begin{remark}[Separation of scales]
The continuous oscillatory decay of PSWF products is independent of
the quadrature-induced defect floor. Without additional structure in
the sampling rule, this decay does not propagate to $E_{mn}$.
\end{remark}

\begin{remark}[Scaling]\label{rem:scaling_stability}
Under DSTP, $\norm{\EN}_2=O(hc)$, so frame stability is governed
directly by $hc/\lambda_{N-1}(c)$, without additional polynomial
growth in $N$.
\end{remark}

%% ==================================================================
\section{Boundary to the companion framework}\label{sec:bridge}
%% ==================================================================

The present paper proves robust finite-dimensional stability under a
standard first-order quadrature consistency assumption, separating
three distinct mechanisms: continuous PSWF orthogonality, oscillatory
decay of the continuous product kernel, and quadrature-induced defect
control.

The continuous bulk--turning analysis of Section~\ref{sec:decay}
proves that off-diagonal PSWF interactions decay analytically in the
bulk sector, and simultaneously demonstrates that this decay is a
purely continuous phenomenon destroyed by generic quadrature.
It is \emph{not used} in the proof of
Theorem~\ref{thm:conditional_frame_schur}: its role is to identify the
structural obstruction that prevents the coarse bound $|E_{mn}|\le Chc$
from exhibiting off-diagonal decay, and thereby to motivate DSTP as
the necessary interface axiom. The unconditional discrete estimate is
governed by $|E_{mn}|\le Chc$; the frame theorem requires the
additional hypothesis DSTP. Consequently the present framework is
robust but not spectrally sharp: it yields frame stability for fixed
$(N,c)$ once the defect is dominated by $\lambda_{N-1}(c)$, but does
not provide an asymptotically uniform theorem in the transition regime
$N\sim c$.

The companion paper~\cite{PaperIIquad} addresses whether oscillatory
structure of PSWF products can be preserved at the discrete level by
PSWF-adapted quadrature. Establishing DSTP for concrete schemes is
deferred entirely to that paper. DSTP is the explicit axiomatic
interface between the two papers. The semiclassical structure underlying
the bulk-sector decay estimates is fully established by Paper~IV
\cite{PaperIV}; the Bulk--Tail programme (Papers~III--IV) and its
connection to the Weil operator are developed in
\cite{PaperII,PaperIV}.

%% ==================================================================
\section{Open problems}\label{sec:open}
%% ==================================================================

\begin{problem}[Sharp decay rate in the bulk sector]
Lemma~\ref{lem:bulk_decay_continuous} establishes algebraic decay
$(1+|m-n|)^{-p}$ for all finite $p$ in the bulk sector.
The question whether this can be upgraded to exponential decay
$e^{-\alpha|m-n|}$ for some $\alpha>0$ in the strict bulk
$m,n\le\gamma_0 N_{\mathrm{Sh}}$ is addressed in Paper~V \cite{PaperV},
which establishes exponential concentration under a WKB stability
condition; we record it here as a sharpness question for the present
paper's IBP approach.
\end{problem}

\begin{problem}[Sharper turning bounds]
Refine the turning contribution from $O(c^{-1/3})$ via Airy asymptotics
and microlocal analysis near the turning points.
\end{problem}

\begin{problem}[Quantitative constants]
Make the constants in Lemmas~\ref{lem:bulk_spectral_sep},
\ref{lem:phase_gap_final}, \ref{lem:ibp_uniform},
and~\ref{lem:derivative_bound} fully explicit in $N$ and $c$.
\end{problem}

\begin{problem}[Oscillation-compatible quadrature]
Identify concrete quadrature schemes satisfying DSTP; a PSWF-Gauss/XRY
framework is developed in~\cite{PaperIIquad}.
\end{problem}

\begin{problem}[Uniform-grid unisolvence]
Determine whether uniform grids are unisolvent for $V_{N,c}$.
\end{problem}

\begin{problem}[Coupled limits]
Analyse $\lambda_{\min}(G^{\mathrm{w}}_{N,c})$ when $N$ and $c$ vary
simultaneously.
\end{problem}

%% ==================================================================
\appendix
\section{Airy uniformisation near turning points}\label{app:airy}
%% ==================================================================

\begin{lemma}[PSWF Airy approximation]\label{lem:airy_approx}
For each $0\le n\le N-1$ and $x\in U_n$,
\[
  \psi_n^{(c)}(x)
  =\kappa_n^{(c)}\,\mathrm{Ai}\!\bigl(-\zeta_n(x)\bigr)+O(c^{-1/3}),
\]
where $\zeta_n(x)=c^{2/3}(x-x_n^*)/T+O(c^{-1/3})$ and
$|\kappa_n^{(c)}|\le C_\kappa$ uniformly in $n,c$
\cite{Olver1974,OsipovRokhlinXiao}.
\end{lemma}
\begin{proof}
Standard semiclassical analysis \cite[Ch.~11]{Olver1974}.
\end{proof}

\begin{lemma}[Turning-point integral via Airy]\label{lem:airy_bound}
For all $0\le m,n\le N-1$,
\[
  \left|\int_{U_m\cup U_n}
    \psi_m^{(c)}(x)\overline{\psi_n^{(c)}(x)}\,dx\right|
  \le C\,c^{-1/3}.
\]
\end{lemma}
\begin{proof}
By the uniform sup-norm bound and $|U_k|=O(c^{-1/3})$,
$\|\psi_k^{(c)}\|_{L^2(U_m\cup U_n)}=O(c^{-1/6})$.
Cauchy--Schwarz on $U_m\cup U_n$ gives $O(c^{-1/3})$.
\end{proof}

%% ==================================================================
\begin{thebibliography}{99}

\bibitem{BonamiKaroui}
A.~Bonami and A.~Karoui,
\textit{Spectral decay of the sinc kernel operator and approximation
with prolate spheroidal wave functions}, arXiv:1012.3881.

\bibitem{BonamiKaroui2014}
A.~Bonami and A.~Karoui,
\textit{Uniform approximation and explicit estimates for the prolate
spheroidal wave functions}, arXiv:1405.3676.

\bibitem{HornJohnson2013}
R.~Horn and C.~Johnson,
\textit{Matrix Analysis}, 2nd ed., Cambridge University Press, 2013.

\bibitem{Olver1974}
F.~W.~J.~Olver,
\textit{Asymptotics and Special Functions}, Academic Press, 1974.

\bibitem{OsipovRokhlinXiao}
A.~Osipov, V.~Rokhlin, and H.~Xiao,
\textit{Prolate Spheroidal Wave Functions of Order Zero},
Springer, 2013.

\bibitem{PaperII}
Anonymous Author,
\emph{Scaling Limits, Trace Formula, and Weil-Operator Connections
for Prolate Sampling Operators}, preprint, 2026.

\bibitem{PaperIIquad}
Anonymous Author,
\emph{Conditional Frame Stability of Prolate Sampling Operators via
PSWF Spectral Projection and XRY Quadrature: A Conditional Implication
Framework for Quadrature Compatibility}, preprint, 2026.

\bibitem{PaperIV}
Anonymous Author,
\emph{Semiclassical Equidistribution of Prolate Spheroidal Wave
Function Densities via Pr\"{u}fer Phase Analysis},
preprint, 2026.

\bibitem{PaperV}
Anonymous Author,
\emph{WKB Stability and Exponential Concentration in the Prolate
Bulk Sector}, preprint, 2026.

\bibitem{Slepian1978}
D.~Slepian,
\textit{Prolate spheroidal wave functions, Fourier analysis, and
uncertainty~-- V},
Bell Syst.\ Tech.\ J.\ \textbf{57} (1978), 1371--1430.

\bibitem{Xiao2001}
H.~Xiao, V.~Rokhlin, and N.~Yarvin,
\textit{Prolate spheroidal wave functions, quadrature, and
interpolation},
Inverse Problems \textbf{17} (2001), 805--838.

\end{thebibliography}

\end{document}
