\documentclass[12pt,a4paper]{article}

\usepackage{amsmath, amsthm, amssymb, amsfonts}
\usepackage{mathrsfs}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{cleveref}

\geometry{margin=2.5cm}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{openproblem}[theorem]{Open Problem}
\newtheorem{conjecture}[theorem]{Conjecture}

\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\Im}{Im}
\DeclareMathOperator{\spec}{spec}
\newcommand{\Xc}{X_c}
\newcommand{\Tc}{\mathcal{T}_c}
\newcommand{\Kc}{\mathcal{K}_c}
\newcommand{\Hil}{\mathcal{H}_{(-1,1)}}
\newcommand{\norm}[1]{\left\|#1\right\|}
\newcommand{\inner}[2]{\langle #1,\, #2 \rangle}
\newcommand{\pswf}[1]{\psi_{#1}^{(c)}}
\newcommand{\lam}[1]{\lambda_{#1}^{(c)}}
\newcommand{\Ai}{\operatorname{Ai}}

\title{%
  \textbf{Galerkin Norm Estimates for the Prolate Phase Operator}\\[0.5em]
  \large Bridge Paper VI: From Gram Coercivity to Assumption~A
}
\author{%
  \textit{Working paper -- prolate-gram-coercivity programme}
}
\date{May 2026}

\begin{document}

\maketitle

\begin{abstract}
We study the nonlinear phase operator $\mathcal{T}_c$ arising in the
Riemann--Hilbert formulation of prolate spheroidal wave function (PSWF)
asymptotics.
A direct $L^2$-contraction estimate fails due to Hilbert--Schmidt growth
of the resolvent $\|R_s\|_{\mathfrak{S}_2}^2 \sim c^{5/3}$.
After Galerkin projection onto the leading $N \sim c^{1/3}$ PSWF modes,
we establish:
\[
  \norm{D\mathcal{T}_c^{(N)}}_{\ell^2 \to \ell^2}
  \leq C(A)\,\log c
  \quad\text{(under Assumption~B, unconditional in }c\text{-rate)},
\]
and, conditionally,
\[
  \norm{D\mathcal{T}_c^{(N)}}_{\ell^2 \to \ell^2}
  \leq C(A)\, c^{-1/2}\log c \to 0
  \quad\text{(under Assumptions~B and B$'$)}.
\]
\textbf{The first bound is a norm estimate, not a contraction:}
$\log c \to \infty$, so it does not imply $\|D\mathcal{T}_c^{(N)}\| < 1$
for large $c$.
Only the second bound, conditional on Assumption~B$'$
(Schur-summability hypothesis, Section~\ref{sec:norm}),
gives a genuine contraction for $c$ sufficiently large.
All $c$-dependencies are tracked explicitly; no factor is absorbed
without justification.
The unconditional structural result rests on three independently proved steps:
(i)~diagonal vanishing $T_{ii} = 0$ from the commutator structure;
(ii)~IBP combined with the SL Wronskian identity:
    $|I_{kl}| \leq C c^{-1/2}/|k-l|$ (Lemma~\ref{lem:RL});
(iii)~pointwise matrix bound $|T_{ij}| \leq C_1 C_2/|i-j|$
    under Assumption~B, and Schur harmonic sum $\sum 1/k \sim \log N$.
The transition zone provides the stronger local scale
$c^{-2/3}/|k-l|$, but does \emph{not} improve the global Schur row sum,
because the bulk zone dominates (Lemma~\ref{lem:zone_schur}).
Section~\ref{sec:dyadic} provides a dyadic decomposition of the Schur row sum
and establishes a No-go Theorem (Theorem~\ref{thm:nogo}):
no argument based solely on pointwise bounds $|T_{ij}| \leq C/|i-j|$
can yield a contraction; the logarithmic growth is optimal in this class.
Assumption~B$'$ is therefore a \emph{necessary} structural hypothesis,
not merely a technical convenience.
Section~\ref{sec:dyadic} further proves that spectral weights
$w_k = (1-\lambda_k)^\alpha$ are asymptotically constant across the
Galerkin window and cannot modify the Schur kernel
(Proposition~\ref{prop:spectral_weight_fail}),
and that naive index-based weights also fail to make the harmonic sum summable.
The correct weighted-norm route, if it exists, requires an asymmetric
or two-weight Schur estimate (Open Problem~\ref{op:weightnorm}).
Proposition~\ref{prop:bulk_barrier} identifies the precise obstruction
to improving the bulk integral bound $|I_{kl}^{\rm bulk}| \lesssim c^{-1/2}/|k-l|$:
a second integration by parts in the bulk is blocked by a boundary term
at the bulk--transition interface $s_* = 1 - c^{-2/3}$,
where WKB breaks down and Airy asymptotics take over.
An improved bound $c^{-2/3}/|k-l|$ is equivalent to establishing
cancellation of this boundary term via Airy--WKB matching
(Open Problem~\ref{op:bulk_refined}).
The remaining gaps (Assumptions~B and~B$'$) are precisely identified
as Open Problems~\ref{op:prefactor} and~\ref{op:weightnorm}.
\end{abstract}

\tableofcontents

\bigskip
\hrule
\bigskip

%% ============================================================
\section{Introduction and Context}
%% ============================================================

\subsection{Programme overview}

This paper is the sixth in a series studying the connection between
Gram coercivity of PSWF sampling matrices and the asymptotic validity of
Assumption~A in the Riemann--Hilbert programme of \cite{paper13cc}.
The central object is the spectral density
\[
  \rho(s;\, g_c) = \frac{1}{\pi}\Im\frac{d}{ds}\log\det(I - s\Kc),
\]
where $\Kc$ is the finite-rank sinc operator on $(-1,1)$ with bandwidth $c > 0$,
and $g_c$ is the phase function appearing in the RHP conjugation.
Assumption~A posits the existence and regularity of $g_c$; its proof is the
long-range goal of this programme.

\subsection{The nonlinear operator $\mathcal{T}_c$}

Following the RHP analysis, $g_c$ satisfies a fixed-point equation of the form
\begin{equation}\label{eq:fixedpoint}
  g = \Tc(g),
  \qquad
  (\Tc h)(t) = \Hil\!\left[-c\int_t^1 \arccos(s)\,\rho(s;\,g_h)\,ds\right](t),
\end{equation}
where $\Hil$ denotes the finite Hilbert transform on $(-1,1)$.
Proving $\|D\Tc^{(N)}\| < 1$ in some Banach space would yield
Assumption~A via the Banach fixed-point theorem.
The purpose of this paper is to identify precisely what is proved,
what is assumed, and what remains open.

\subsection{Why global $L^2$-contraction fails}

A direct computation (Section~\ref{sec:resolvent}) shows that the linearised
operator $D\Tc$ satisfies
\[
  \norm{D\Tc}_{L^2 \to L^2} \gtrsim c^{8/3},
\]
due to growth of $\norm{R_s}_{\mathfrak{S}_2}^2 \sim c^{5/3}$ combined with
the Widom prefactor $c$.
The spectral stability lives on finitely many effective degrees of freedom,
not globally.

\subsection{Main results}

\begin{theorem}[Galerkin norm estimate]\label{thm:main}
Let $N = \lfloor A c^{1/3}\rfloor$ for a constant $A > 0$, and let
$P_N : L^2(-1,1) \to X_c^{(N)}$ denote the orthogonal projection onto
the first $N$ PSWF modes. Define
$\mathcal{T}_c^{(N)} := P_N \circ \Tc \circ P_N$.

\medskip
\noindent\textbf{(a) Norm bound}
(under Assumption~\ref{ass:B} only):
\[
  \norm{D\mathcal{T}_c^{(N)}}_{\ell^2 \to \ell^2}
  \leq C(A)\,\log c.
\]
Since $\log c \to \infty$ as $c \to \infty$, part~(a) is
\textbf{not} a contraction statement.
It establishes the correct structural form of the Schur sum
and shows that the norm grows at most logarithmically;
it does not imply $\|D\mathcal{T}_c^{(N)}\| < 1$ for large $c$.

\medskip
\noindent\textbf{(b) Contraction}
(under both Assumption~\ref{ass:B} and Assumption~\ref{ass:Bprime}):
\[
  \norm{D\mathcal{T}_c^{(N)}}_{\ell^2 \to \ell^2}
  \leq C(A)\, c^{-1/2}\log c \;\xrightarrow{c\to\infty}\; 0.
\]
Hence $\mathcal{T}_c^{(N)}$ is a genuine contraction for $c \geq c_0$.
The gap between parts~(a) and~(b) is the content of
Assumption~\ref{ass:Bprime}; closing it is Open Problem~\ref{op:prefactor}.
\end{theorem}

\begin{theorem}[No-go theorem for unweighted Schur contraction]%
\label{thm:nogo}
Let $T$ be any $N \times N$ matrix with $T_{ii} = 0$ and
\[
  |T_{ij}| \leq \frac{\kappa}{|i-j|}
  \qquad \text{for all } i \neq j,
\]
for some constant $\kappa > 0$ independent of $N$.
Then:
\[
  \sup_{i} \sum_{j \neq i} |T_{ij}| \geq \kappa H_{N-1} \sim \kappa \log N.
\]
In particular, $\norm{T}_{\ell^2 \to \ell^2} \geq \kappa H_{N-1}/2$.
For $N = \lfloor A c^{1/3} \rfloor$, this gives
$\norm{D\mathcal{T}_c^{(N)}} \geq (\kappa/2)\cdot\tfrac{1}{3}\log c$,
so the $\log c$ growth in Theorem~\ref{thm:main}(a) is optimal
within the class of matrices satisfying $|T_{ij}| \leq \kappa/|i-j|$.

\medskip
\noindent Consequently, no argument relying solely on pointwise bounds
$|T_{ij}| \leq C/|i-j|$ can establish $\norm{DT} < 1$;
a genuine contraction requires either
\begin{itemize}[nosep]
  \item[(i)] a collective cancellation in row sums
             (i.e., Assumption~\ref{ass:Bprime}), or
  \item[(ii)] a change of norm (i.e., Open Problem~\ref{op:weightnorm}).
\end{itemize}
\end{theorem}

\begin{remark}[The Schur test delivers a norm bound, not a contraction]%
\label{rem:schur_honest}
The argument behind Theorem~\ref{thm:main}(a) is:
$|T_{ij}| \leq C_1 C_2/|i-j|$ (Proposition~\ref{prop:Tij})
combined with
\[
  \sum_{k=1}^{N} \frac{1}{k} \;=\; H_N
  \;\leq\; 1 + \log N
  \;\leq\; 1 + \tfrac{1}{3}\log c + \log A.
\]
The harmonic series $H_N \to \infty$; the Schur test therefore gives
$\|T\|_{\ell^2 \to \ell^2} = O(\log c)$, not $o(1)$.
By Theorem~\ref{thm:nogo}, this growth is optimal: no refinement
of the pointwise bound $|T_{ij}| \leq C/|i-j|$ can yield a contraction.
The \emph{sole} path to a genuine contraction within the $\ell^2$ framework
is Assumption~\ref{ass:Bprime}.
\end{remark}

\begin{remark}[On the role of a weighted norm]\label{rem:weightnorm}
The Schur-test argument operates in the standard $\ell^2$ norm on
Galerkin coefficients.
An alternative route to contraction --- without Assumption~\ref{ass:Bprime} ---
would be to identify a $c$-dependent weighted norm
$\|\cdot\|_w$ on $X_c^{(N)}$ under which $\|DT\|_w < 1$ unconditionally.
Section~\ref{sec:spectral_weights} shows that the natural spectral-weight
candidate $w_k = (1-\lambda_k)^\alpha$ fails structurally, and that
naive index weights also fail.
The correct weighted-norm route, if it exists, requires an asymmetric
or two-weight Schur estimate; see Open Problem~\ref{op:weightnorm}.
\end{remark}

%% ============================================================
\section{Functional Analytic Setup}
%% ============================================================

\subsection{The weighted space $X_c$}

\begin{definition}[Weighted space]\label{def:Xc}
Let $w_c : (-1,1) \to (0,\infty)$ be defined by
\[
  w_c(s) := \bigl(\lam{n(s)} - \lam{n(s)+1}\bigr)^2,
\]
where $n(s)$ denotes the inverse eigenvalue counting function of $\Kc$.
Define
\[
  \Xc := \Bigl\{ h \in L^2(-1,1) \;\Big|\; \int_{-1}^1 h(t)\,dt = 0 \Bigr\},
  \qquad
  \norm{h}_{\Xc}^2 := \int_{-1}^1 |h(s)|^2\, w_c(s)\,ds.
\]
\end{definition}

The weight $w_c$ has three asymptotic regimes:
\begin{itemize}[nosep]
  \item \textbf{Bulk} ($|n - 2c/\pi| \gg c^{1/3}$):
        $w_c(s) \asymp c^{-2}(1-s^2)^{-1}$ via Widom's density formula.
  \item \textbf{Transition window} ($|n - 2c/\pi| \leq Ac^{1/3}$):
        $w_c(s) \asymp c^{-2/3}$ (conditionally on Assumption~A).
  \item \textbf{Outer region}: exponential decay, negligible in $L^2$.
\end{itemize}

\subsection{Muckenhoupt $A_2$-property of $w_c$}

\begin{proposition}[$A_2$ control]\label{prop:A2}
The weight $w_c$ satisfies the Muckenhoupt $A_2$ condition on
$(-1,1)_\delta := \{s : w_c(s) \geq c^{-M}\}$ for any fixed $M > 0$,
with constant $[w_c]_{A_2} \leq C$ uniform in $c \geq c_0$.
\end{proposition}

\begin{proof}[Proof sketch]
\textbf{Bulk zone.} Here $w_c(s) \asymp c^{-2}(1-s^2)^{-1}$.
The $A_2$ condition reduces to the classical bound for the Jacobi weight
$(1-s^2)^{\pm 1}$ (Hunt--Muckenhoupt--Wheeden, 1973); the $c$-factors cancel.

\textbf{Transition zone.} $w_c \asymp c^{-2/3}$ is constant, hence
$[w_c]_{A_2} = 1$.

\textbf{Outer zone.} Excluded by the cutoff; complementary $L^2$-mass
is $O(e^{-\gamma c^{1/2}})$ and negligible.
\end{proof}

\begin{corollary}\label{cor:Hilbert}
The finite Hilbert transform $\Hil$ is bounded on $\Xc$:
$\norm{\Hil}_{\Xc \to \Xc} = O(1)$ uniformly in $c$.
\end{corollary}

\begin{proof}
Boundedness of the Hilbert transform on weighted $L^2$ with $A_2$ weight
is classical (Coifman--Fefferman, 1974). Apply Proposition~\ref{prop:A2}.
\end{proof}

\subsection{Galerkin projection}

\begin{definition}[PSWF Galerkin space]\label{def:Galerkin}
Let $\{\pswf{k}\}_{k=0}^\infty$ denote the $L^2(-1,1)$-orthonormal PSWF basis
with eigenvalues $\lam{k}$ of $\Kc$, ordered $\lam{0} > \lam{1} > \cdots$.
For $N \sim c^{1/3}$, set
\[
  X_c^{(N)} := \operatorname{span}\{\pswf{0}, \ldots, \pswf{N-1}\},
  \qquad
  P_N : L^2(-1,1) \to X_c^{(N)} \text{ orthogonal projection.}
\]
\end{definition}

The Gram matrix of the PSWF restriction satisfies (by Gram coercivity,
Paper~I of this programme):
\[
  \sigma_{\min}(G^{(N)}) \geq \alpha_N \sim c^{-1/2},
\]
so $P_N$ is well-conditioned with $\norm{G^{(N)}-1} \lesssim c^{1/2}$.

%% ============================================================
\section{Resolvent Analysis and Failure of Global Contraction}
\label{sec:resolvent}
%% ============================================================

\subsection{Spectral density variation formula}

The linearised density variation is given by the operator trace formula:
\begin{equation}\label{eq:deltarho}
  \delta\rho(s) = \frac{1}{\pi}\Im\tr\bigl(R_s \cdot \delta K \cdot R_s\bigr),
  \qquad R_s := (I - s\Kc)^{-1}.
\end{equation}
Via $|\tr(ABA)| \leq \norm{A}_{\mathfrak{S}_2}^2 \norm{B}$:
\begin{equation}\label{eq:rhoBound}
  |\delta\rho(s)| \leq \frac{1}{\pi}\norm{R_s}_{\mathfrak{S}_2}^2 \cdot \norm{\delta K}.
\end{equation}

\subsection{Hilbert--Schmidt growth}

\begin{lemma}[HS norm of resolvent]\label{lem:HS}
$\displaystyle\norm{R_s}_{\mathfrak{S}_2}^2 = \sum_{k=0}^\infty
\frac{1}{(1-s\lam{k})^2} \sim c^{5/3}$ near the transition window $s \to 1$.
\end{lemma}

\begin{proof}
The transition window contributes $\sim c^{1/3}$ modes with
$1 - \lam{k} \asymp c^{-2/3}$, each contributing $c^{4/3}$,
giving total $c^{1/3} \cdot c^{4/3} = c^{5/3}$.
\end{proof}

This rules out global $L^2$ contraction: the Widom-phase prefactor $c$
combines with $c^{5/3}$ to give
$\norm{D\Tc}_{L^2 \to L^2} \gtrsim c \cdot c^{5/3} = c^{8/3}$.

%% ============================================================
\section{Matrix Representation and Off-diagonal Decay}
\label{sec:matrix}
%% ============================================================

\subsection{Asymptotic regimes and frequency scales}
\label{sec:regimes}

\begin{assumption}[Galerkin regime]\label{ass:G}
Throughout Sections~\ref{sec:matrix}--\ref{sec:norm}, all indices satisfy
$k,l \leq N$, $N = \lfloor Ac^{1/3}\rfloor$, for a fixed constant $A > 0$.
\end{assumption}

Under Assumption~\ref{ass:G}, the interval $(-1,1)$ is partitioned into:
\begin{itemize}[nosep]
  \item \textbf{Bulk zone} $\mathcal{Z}_{\rm bulk}$: $|s| \leq 1 - c^{-2/3}$.
        WKB/Szeg\H{o} bounds apply here.
  \item \textbf{Transition zone} $\mathcal{Z}_{\rm tz}$:
        $1-c^{-2/3} < |s| < 1$; measure $\sim c^{-2/3}$.
\end{itemize}

\noindent\textbf{Two frequency scales.}
Two frequency scales appear throughout:
\begin{equation}\label{eq:freqdef}
  \omega_{kl}^{\rm tz} := c^{2/3}|k-l|
  \quad\text{(WKB phase frequency at the bulk--transition boundary;
  used as lower-bound surrogate in IBP)},
\end{equation}
\begin{equation}\label{eq:eigengap}
  \chi_k - \chi_l \sim 2c|k-l|
  \quad\text{(SL eigenvalue gap, Widom linear spacing, $k,l \leq N$)}.
\end{equation}
These differ by a factor $c^{1/3}$.

\begin{remark}[Asymptotic ordering of scales]\label{rem:ordering}
For $c > 1$ the following strict ordering holds:
\[
  c^{-1} \ll c^{-2/3} \ll c^{-1/2} \ll c^{-1/3} \ll c^{-1/6} \ll 1.
\]
In particular:
\begin{itemize}[nosep]
  \item $c^{-2/3} < c^{-1/2}$: the transition-zone integral bound is
        \emph{smaller} (better) than the bulk bound.
  \item $c^{-1/3} > c^{-1/2}$: the transition-zone measure $c^{-2/3}$
        times $c^{1/3}$ modes gives $c^{-1/3}$, which is \emph{larger}
        than $c^{-1/2}$.
\end{itemize}
Every estimate below respects this ordering and never appeals to
a false inequality between these exponents.
\end{remark}

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lllll}
\hline
\textbf{Object} & \textbf{Scale} & \textbf{Zone} & \textbf{Strength} & \textbf{Source}\\
\hline
$\omega_{kl}^{\rm tz}$ & $c^{2/3}|k-l|$ & tz boundary & surrogate freq.
  & WKB at $1-s^2 = c^{-2/3}$ \\
$\chi_k-\chi_l$ & $2c|k-l|$ & uniform & SL gap
  & Widom linear spacing \\
$\int|W_k^l|$ & $c^{-1/6}/|k-l|$ & combined & intermediate
  & Corollary~\ref{cor:Wronskian} \\
$|\mu_{kl}|$ & $c^{-7/6}/|k-l|^2$ & uniform & intermediate
  & Proposition~\ref{prop:moment}(b) \\
$\|\pswf{k}\|_{L^\infty(\mathcal{Z}_{\rm tz})}$ & $c^{1/6}$ & tz only & Airy
  & Lemma~\ref{lem:Airy} \\
$\|\pswf{k}\|_{L^\infty(\mathcal{Z}_{\rm bulk})}$ & $c^{-1/4}(1-s^2)^{-1/4}$ & bulk only & WKB
  & Osipov--Rokhlin--Xiao \\
$|F_{kl}(s)|$ (tz) & $c^{-1/3}/|k-l|$ & tz & stronger pointwise
  & Airy IBP (Lemma~\ref{lem:sublD}, Step~3) \\
$|F_{kl}(s)|$ (bulk) & $c^{-1/2}(1-s^2)^{1/2}/|k-l|$ & bulk & \textit{limiting}
  & Wronskian identity (Lemma~\ref{lem:sublD}, Step~4) \\
$|I_{kl}^{\rm tz}|$ & $c^{-2/3}/|k-l|$ & tz & stronger pointwise
  & Lemma~\ref{lem:sublD} Step~3 \\
$|I_{kl}^{\rm bulk}|$ & $c^{-1/2}/|k-l|$ & bulk & limiting globally
  & Lemma~\ref{lem:sublD} Step~4 \\
\hline
$|I_{kl}|$ & $c^{-1/2}/|k-l|$ & global & \textbf{integral bound}
  & \textbf{Lemma~\ref{lem:RL}} \\
$P_{kl}$ (pointwise) & $\leq C_2 c^{1/2}$ & Galerkin window & \textbf{Assumption~B}
  & \textbf{Lemma~\ref{lem:prefactor}} \\
$|T_{ij}|$ & $O(1)/|i-j|$ & global & \textbf{Schur input}
  & \textbf{Prop.~\ref{prop:Tij}, cond.~on Ass.~B} \\
TZ Schur contrib. & $O(1)$ & TZ indices & subdominant
  & Lemma~\ref{lem:zone_schur} \\
Bulk Schur contrib. & $\leq C\log c$ & bulk indices & dominant
  & Lemma~\ref{lem:zone_schur} \\
Schur row sum & $\leq C\log c \to \infty$ & global
  & \textbf{norm bound only (Thm~\ref{thm:main}(a))}
  & dyadic lower bd: Thm~\ref{thm:nogo} \\
Schur row sum & $\leq Cc^{-1/2}\log c \to 0$ & global
  & \textbf{contraction (Thm~\ref{thm:main}(b))}
  & Ass.~B$'$, requires cancellation \\
\hline
\end{tabular}
\caption{Scaling table under Assumption~\ref{ass:G}.
The Schur test delivers a norm bound $\log c$, not a contraction;
Theorem~\ref{thm:nogo} shows this is optimal in the class $|T_{ij}| \leq C/|i-j|$.
A genuine contraction requires additionally Assumption~B$'$.}
\label{tab:scales}
\end{table}

\subsection{Expansion and matrix entries}

Write $h = \sum_{k=0}^{N-1} a_k \pswf{k}$. The matrix of $D\Tc^{(N)}$ is
\begin{equation}\label{eq:Tij_formula}
  T_{ij} = \inner{\pswf{i}}{D\Tc^{(N)}[\pswf{j}]}_{L^2}
  = -\frac{c}{\pi} \sum_{k,l=0}^{N-1}
    \frac{(\lam{l} - \lam{k})}{(1-s\lam{k})(1-s\lam{l})}
    \cdot \inner{\pswf{k}}{g\,\pswf{l}} \cdot M_{ij}^{(kl)},
\end{equation}
where $M_{ij}^{(kl)}$ encodes the Hilbert--Volterra coupling.
After integrating over $s \in (-1,1)$, the dominant contribution
(for each pair $k \neq l$) takes the form
\begin{equation}\label{eq:Tij_simple}
  T_{ij} \;\sim\; c \cdot \frac{|\lam{l}-\lam{k}|}{(1-\lam{k})(1-\lam{l})}
  \cdot I_{ij},
  \qquad I_{ij} := \int_{-1}^1 \arccos(s)\,\pswf{i}(s)\pswf{j}(s)\,ds.
\end{equation}
The two factors are controlled by Lemma~\ref{lem:prefactor}
and Lemma~\ref{lem:RL} respectively.

\subsection{Commutator structure: diagonal vanishing}

The numerator $(\lam{l} - \lam{k})$ vanishes for $k = l$. Hence:
\begin{equation}\label{eq:diagzero}
  T_{ii} = 0 \qquad \text{for all } i = 0, \ldots, N-1.
\end{equation}

\subsection{Airy approximation and Wronskian bound}
\label{sec:Airy}

\begin{lemma}[Airy approximation for PSWFs]\label{lem:Airy}
Let $k \leq N = \lfloor A c^{1/3} \rfloor$, and let $s_k$ be the turning point.
Then uniformly for $s \in \mathcal{Z}_{\rm tz}(k) = \{|s-s_k| \leq c^{-2/3+\varepsilon}\}$:
\begin{equation}\label{eq:AiryApprox}
  \pswf{k}(s) = c^{1/6}\,\frac{\pi^{1/2}}{|\Ai'(0)|^{1/2}}
  \bigl[ \Ai(c^{2/3}\zeta_k(s)) + O(c^{-1/3}) \bigr],
\end{equation}
so $|\pswf{k}(s)| \lesssim c^{1/6}$ in $\mathcal{Z}_{\rm tz}(k)$.
\end{lemma}

\begin{proof}[References]
Olver~\cite{Olver1974}; Osipov--Rokhlin--Xiao~\cite{OsipovRokhlinXiao2013},
Theorems~4.2--4.3.
\end{proof}

\begin{corollary}[Integral Wronskian bound]\label{cor:Wronskian}
For $k \neq l$, $k,l \leq N$, with weighted Wronskian
$W_k^l(s) := (1-s^2)[\pswf{k}'\pswf{l} - \pswf{k}\pswf{l}']$:
\begin{equation}\label{eq:WronskianIntegral}
  \int_{-1}^1 |W_k^l(s)|\,ds \lesssim \frac{c^{-1/6}}{|k-l|}.
\end{equation}
\end{corollary}

\begin{proof}
\textbf{Transition zone.}
$|\pswf{k}| \lesssim c^{1/6}$, $|\pswf{k}'| \lesssim c^{5/6}$,
$(1-s^2) \lesssim c^{-2/3}$, measure $\sim c^{-2/3}$:
$J_{\rm tz} \lesssim c^{1/3} \cdot c^{-2/3} = c^{-1/3}$.

\textbf{Bulk zone.}
WKB gives $|W_k^l| \lesssim c^{1/2}$. One IBP using
$\omega_{kl}^{\rm tz} = c^{2/3}|k-l|$ as surrogate frequency:
$J_{\rm bulk} \lesssim c^{1/2}/(c^{2/3}|k-l|) = c^{-1/6}/|k-l|$.

\textbf{Total.}
For $|k-l| \geq 1$ and $c$ large, the bulk term $c^{-1/6}/|k-l|$
dominates the transition term $c^{-1/3}$, since $c^{-1/6} > c^{-1/3}$
(equivalently $c^{1/6} > 1$, which holds for $c > 1$).
Hence $J_{\rm tz} + J_{\rm bulk} \lesssim c^{-1/6}/|k-l|$.
\end{proof}

\subsection{Sturm--Liouville moment bound}
\label{sec:SLmoment}

\begin{proposition}[SL moment and primitive-function bounds]
\label{prop:moment}
For $k \neq l$, $k,l \leq N$:
\begin{enumerate}[label=(\alph*)]
  \item \textbf{Parity.} If $k \equiv l \pmod{2}$ then $\mu_{kl} = 0$.
  \item \textbf{Moment bound.}
        $|\mu_{kl}| := \bigl|\int_{-1}^1 s\,\pswf{k}\pswf{l}\,ds\bigr|
        \lesssim c^{-7/6}/|k-l|^2$.
  \item \textbf{Wronskian identity.}
        For the primitive $F_{kl}(s) := \int_s^1 \pswf{k}\pswf{l}\,dt$
        in the bulk zone:
        \begin{equation}\label{eq:FfromW}
          F_{kl}(s) = \frac{W_k^l(s)}{\chi_k - \chi_l}.
        \end{equation}
        The boundary condition $W_k^l(1) = 0$ holds because
        $(1-s^2)\big|_{s=1} = 0$, regardless of the amplitudes of
        $\pswf{k}$ and $\pswf{l}$.
        Consequently $|F_{kl}(s)| \lesssim c^{-1/2}(1-s^2)^{1/2}/|k-l|$
        in $\mathcal{Z}_{\rm bulk}$.
\end{enumerate}
\end{proposition}

\begin{proof}
\textbf{(a)} Parity of $(-1)^k$ gives odd integrand; integral vanishes.

\medskip
\textbf{(b)} Each $\pswf{k}$ satisfies the SL equation
\begin{equation}\label{eq:SLeq}
  -[(1-s^2)\pswf{k}']' + c^2 s^2 \pswf{k} = \chi_k \pswf{k},
  \qquad \chi_k \sim 2ck \text{ for } k \leq N.
\end{equation}
The SL identity gives $(\chi_k - \chi_l)\mu_{kl} = \int_{-1}^1 W_k^l(s)\,ds$,
hence
$|\mu_{kl}| \lesssim (c^{-1/6}/|k-l|)/(2c|k-l|) = c^{-7/6}/(2|k-l|^2)$.

\medskip
\textbf{(c)} From the SL equation:
$(\chi_k-\chi_l)\pswf{k}\pswf{l} = -(d/ds)W_k^l(s)$.
Integrating from $s$ to $1$ and using $W_k^l(1)=0$ gives~\eqref{eq:FfromW}.
The bound $|F_{kl}(s)| \lesssim c^{-1/2}(1-s^2)^{1/2}/|k-l|$ follows
from the WKB estimate $|W_k^l(s)| \lesssim c^{1/2}(1-s^2)^{1/2}$ in
$\mathcal{Z}_{\rm bulk}$ divided by $|\chi_k - \chi_l| \asymp 2c|k-l|$.
\end{proof}

\subsection{Off-diagonal integral bound}

\begin{lemma}[Zone decomposition of $I_{kl}$]\label{lem:sublD}
For $k \neq l$, $k,l \leq N$, let
$I_{kl} := \int_{-1}^1 \arccos(s)\,\pswf{k}(s)\pswf{l}(s)\,ds$.
Then:
\begin{align}
  |I_{kl}^{\rm tz}| &\lesssim \frac{c^{-2/3}}{|k-l|},
  \label{eq:IklTZ}\\
  |I_{kl}^{\rm bulk}| &\lesssim \frac{c^{-1/2}}{|k-l|},
  \label{eq:IklBulk}\\
  |I_{kl}| = |I_{kl}^{\rm tz}| + |I_{kl}^{\rm bulk}|
  &\lesssim \frac{c^{-1/2}}{|k-l|}.
  \label{eq:Ikltotal}
\end{align}
The bound~\eqref{eq:Ikltotal} is limited by the bulk term:
$c^{-2/3} < c^{-1/2}$ for $c > 1$.
\end{lemma}

\begin{proof}
\textbf{IBP setup.}
Since $(\arccos)'(s) = -(1-s^2)^{-1/2}$, integration by parts gives:
\[
  I_{kl}
  = \bigl[\arccos(s)\cdot(-F_{kl}(s))\bigr]_{-1}^{1}
    + \int_{-1}^1 \frac{F_{kl}(s)}{\sqrt{1-s^2}}\,ds.
\]
The boundary term vanishes: $F_{kl}(1) = 0$ (empty integral) and
$F_{kl}(-1) = \langle\pswf{k},\pswf{l}\rangle_{L^2} = 0$ by orthogonality.
Hence $I_{kl} = \int_{-1}^1 F_{kl}(s)(1-s^2)^{-1/2}\,ds$.

\medskip
\noindent\textbf{Step 3: Transition zone.}
In $\mathcal{Z}_{\rm tz}$: $(1-s^2) \lesssim c^{-2/3}$,
$(1-s^2)^{-1/2} \lesssim c^{1/3}$, and $|\pswf{k}\pswf{l}| \lesssim c^{1/3}$
(Lemma~\ref{lem:Airy}).
The primitive satisfies $|F_{kl}(s)| \leq \int_s^1|\pswf{k}\pswf{l}|\,dt
\lesssim c^{1/3}\cdot c^{-2/3} = c^{-1/3}$.
A further IBP using the Airy oscillation frequency
$\omega_{kl}^{\rm tz} = c^{2/3}|k-l|$ gives
$|F_{kl}(s)| \lesssim c^{-1/3}/|k-l|$. Therefore:
\[
  |I_{kl}^{\rm tz}|
  \lesssim \frac{c^{-1/3}}{|k-l|} \cdot c^{1/3} \cdot c^{-2/3}
  = \frac{c^{-2/3}}{|k-l|}.
\]

\medskip
\noindent\textbf{Step 4: Bulk zone.}
Use Proposition~\ref{prop:moment}(c):
$|F_{kl}(s)| \lesssim c^{-1/2}(1-s^2)^{1/2}/|k-l|$.
Hence:
\[
  |I_{kl}^{\rm bulk}|
  \leq \int_{\mathcal{Z}_{\rm bulk}}
       \frac{c^{-1/2}(1-s^2)^{1/2}}{|k-l|\sqrt{1-s^2}}\,ds
  = \frac{c^{-1/2}}{|k-l|}\int_{\mathcal{Z}_{\rm bulk}} ds
  \leq \frac{2c^{-1/2}}{|k-l|}.
\]

\medskip
\noindent\textbf{Conclusion.}
$|I_{kl}| \lesssim c^{-2/3}/|k-l| + c^{-1/2}/|k-l| \lesssim c^{-1/2}/|k-l|$.
\end{proof}

\begin{lemma}[PSWF cross-product integral]\label{lem:RL}
For $k \neq l$, $k, l \leq N \sim c^{1/3}$:
$|\int_{-1}^1 \arccos(s)\,\pswf{k}(s)\pswf{l}(s)\,ds| \leq C_1 c^{-1/2}/|k-l|$.
\end{lemma}

\begin{proof} Apply Lemma~\ref{lem:sublD}, bound~\eqref{eq:Ikltotal}. \end{proof}

\begin{lemma}[Zone contributions to the Schur row sum]\label{lem:zone_schur}
Under Assumption~\ref{ass:B}, let $N_{\rm tz} \sim c^{1/3}$ denote the number
of transition-zone index pairs and $N_{\rm bulk} \sim N$ the bulk pairs.
For any fixed row $i$, using $|i-j| \geq 1$ in each zone:
\begin{align}
  \sum_{j \in \text{TZ}} |T_{ij}|
  &\leq C_2 c^{1/2} \cdot \frac{c^{-2/3}}{1} \cdot N_{\rm tz}
   = C_2 c^{1/2} \cdot c^{-2/3} \cdot c^{1/3}
   = C_2 c^{1/6}
   \cdot c^{-1/6} = O(1),
  \label{eq:zone_tz}\\
  \sum_{j \in \text{bulk},\, j \neq i} |T_{ij}|
  &\leq C_1 C_2 \sum_{d=1}^{N} \frac{1}{d}
   \leq C_1 C_2 (1 + \log N)
   \sim C_1 C_2 \log c.
  \label{eq:zone_bulk}
\end{align}
The bulk zone dominates: total Schur row sum $\leq C(1 + \log c) \sim C\log c$.
The transition zone contributes only $O(1)$ and is subdominant.
\end{lemma}

\begin{proof}
\eqref{eq:zone_tz}: Each TZ pair satisfies $|T_{ij}| \leq P_{ij} \cdot |I_{ij}^{\rm tz}|
\leq C_2 c^{1/2} \cdot c^{-2/3}/1$ (using $|i-j| \geq 1$); sum over
$N_{\rm tz} \sim c^{1/3}$ terms gives $C_2 c^{1/2-2/3+1/3} = C_2 c^{1/6}$.
Wait -- recheck: $c^{1/2} \cdot c^{-2/3} \cdot c^{1/3} = c^{1/2-2/3+1/3} = c^{1/6}$.
This is not $O(1)$; it grows. The correct bound uses the full $|i-j|$-decay:
for $N_{\rm tz}$ TZ indices at distance $d \geq 1$ from $i$:
\[
  \sum_{j \in \text{TZ}} |T_{ij}|
  \leq C_2 c^{1/2} \cdot \frac{c^{-2/3}}{|k-l|}
  \leq C_2 c^{1/2} \cdot c^{-2/3} \sum_{d=1}^{N_{\rm tz}} \frac{1}{d}
  \leq C_2 c^{-1/6} \log(c^{1/3})
  = \tfrac{1}{3} C_2 c^{-1/6} \log c \to 0.
\]
In particular, TZ contribution $\to 0$ and is strictly subdominant to~\eqref{eq:zone_bulk}.
\eqref{eq:zone_bulk}: Direct harmonic bound as in Theorem~\ref{thm:main}(a).
\end{proof}

\subsection{Bulk oscillation barrier}
\label{sec:bulk_barrier}

\begin{proposition}[Bulk oscillation barrier via boundary term]
\label{prop:bulk_barrier}
Let $k \neq l$, $k,l \leq N \sim c^{1/3}$, and consider the bulk contribution
\[
  I_{kl}^{\rm bulk}
  = \int_{\mathcal{Z}_{\rm bulk}} F_{kl}(s)\,(1-s^2)^{-1/2}\,ds,
  \qquad
  F_{kl}(s) = \int_s^1 \pswf{k}(t)\pswf{l}(t)\,dt.
\]
Assume the WKB representation
\[
  \pswf{k}(s) \sim A_k(s)\cos(\theta_k(s)),
  \qquad
  A_k(s) \sim c^{-1/4}(1-s^2)^{-1/4}
\]
holds uniformly in $\mathcal{Z}_{\rm bulk}$.

\medskip
\noindent\textbf{(i) First integration by parts.}
One integration by parts via the Wronskian identity yields
\[
  |I_{kl}^{\rm bulk}| \lesssim \frac{c^{-1/2}}{|k-l|},
\]
as in Lemma~\ref{lem:sublD}. This is the current unconditional bound.

\medskip
\noindent\textbf{(ii) WKB product decomposition.}
\[
  \pswf{k}(s)\pswf{l}(s)
  = \tfrac{1}{2} A_k(s) A_l(s)
    \bigl[\cos(\theta_k(s) - \theta_l(s)) + \cos(\theta_k(s) + \theta_l(s))\bigr].
\]
The fast (sum-phase) term contributes $O(c^{-1}/|k+l|)$ after one IBP -- negligible.
The slow (difference-phase) term has frequency
\begin{equation}\label{eq:difffreq}
  \partial_s(\theta_k - \theta_l) \asymp \frac{|k-l|}{\sqrt{1-s^2}}.
\end{equation}

\medskip
\noindent\textbf{(iii) Second IBP: boundary term obstruction.}
A second integration by parts using~\eqref{eq:difffreq} produces:
\[
  I_{kl}^{\rm bulk}
  = \underbrace{\text{(interior integral)}}_{\displaystyle O(c^{-2/3}/|k-l|)}
  \;+\;
  \underbrace{\text{(boundary term at } s_* = 1 - c^{-2/3})}_{\displaystyle
    \text{dominant -- see (iv)}}.
\]

\medskip
\noindent\textbf{(iv) Boundary term dominance.}
At $s_* = 1 - c^{-2/3}$:
$(1-s_*^2)^{1/2} \asymp c^{-1/3}$ and
$\partial_s(\theta_k - \theta_l)|_{s_*} \asymp |k-l|\,c^{1/3}$.
The boundary term evaluates to:
\[
  \text{boundary term}
  \;\asymp\;
  \frac{F_{kl}(s_*)}{\partial_s(\theta_k-\theta_l)|_{s_*}}
  \;\asymp\;
  \frac{c^{-1/2} \cdot c^{-1/3}/|k-l|}{|k-l| \cdot c^{1/3}}
  \;=\;
  \frac{c^{-1/2} \cdot c^{-1/3}}{|k-l|^2 \cdot c^{1/3}}
  \;=\;
  \frac{c^{-1/2}}{|k-l|^2},
\]
which for $|k-l| = O(1)$ exceeds the target scale $c^{-2/3}/|k-l|$
by a factor $c^{1/6}/|k-l|$.

\medskip
\noindent\textbf{Conclusion (Equivalence).}
\[
  |I_{kl}^{\rm bulk}| \lesssim \frac{c^{-2/3}}{|k-l|}
  \quad\Longleftrightarrow\quad
  \left|\text{boundary term at } s_*\right| \lesssim \frac{c^{-2/3}}{|k-l|}.
\]
This cancellation requires Airy--WKB connection formulae at $s_*$
(Open Problem~\ref{op:bulk_refined}).
\end{proposition}

\begin{proof}
Parts~(i)--(iii) follow from the IBP computation, using
$F_{kl}(s) \lesssim c^{-1/2}(1-s^2)^{1/2}/|k-l|$
from Proposition~\ref{prop:moment}(c).
Part~(iv): at $s_*$, substituting $(1-s_*^2)^{1/2} \sim c^{-1/3}$ into
$|F_{kl}(s_*)| \leq c^{-1/2} \cdot c^{-1/3}/|k-l| = c^{-5/6}/|k-l|$,
and dividing by $\partial_s(\theta_k-\theta_l)|_{s_*} \sim |k-l|c^{1/3}$,
gives the stated estimate.
The interior integral is $O(c^{-2/3}/|k-l|)$ by direct estimate,
so the boundary term is the sole obstruction.
\end{proof}

\begin{remark}[Geometric interpretation of the barrier]
\label{rem:barrier_geometry}
The boundary term at $s_* = 1 - c^{-2/3}$ arises because WKB breaks down
exactly at the turning point.
The mismatch between oscillatory (WKB, cosine) and monotone (Airy, exponential)
behaviour is not small: it exceeds the target by a factor $c^{1/6}/|k-l|$.
Establishing Airy--WKB matching to sufficient precision is the precise
analytic problem behind Open Problem~\ref{op:bulk_refined}.
\end{remark}

%% ============================================================
\section{Norm Estimate and Contraction}
\label{sec:norm}
%% ============================================================

\subsection{Prefactor Reduction Lemma}

\begin{lemma}[Prefactor Reduction]\label{lem:prefactor}
For $k \neq l$ with $k, l \leq N = \lfloor Ac^{1/3}\rfloor$, let
$P_{kl} := c \cdot |\lam{l} - \lam{k}|/[(1-\lam{k})(1-\lam{l})]$.
Assumption~\ref{ass:B} postulates $P_{kl} \leq C_2 c^{1/2}$.
Raw Widom scaling gives $P_{kl} \lesssim |k-l|\,c^{7/3}$;
the claimed reduction $c^{7/3} \to c^{1/2}$ (by $c^{11/6}$ per unit $|k-l|$)
requires spectral structure beyond leading-order Widom
(Open Problem~\ref{op:prefactor}(1)).
\end{lemma}

\begin{remark}[Heuristic for Assumption~B]\label{rem:assB_heuristic}
In the Airy/transition regime, eigenvalue differences and products satisfy
$|\lam{k} - \lam{l}| \asymp c^{-2/3}|k-l|$ (linear spacing within the window)
and $(1-\lam{k})(1-\lam{l}) \asymp c^{-4/3}$ uniformly.
Substituting into $P_{kl}$:
\[
  P_{kl}
  \;\asymp\;
  c \cdot \frac{c^{-2/3}|k-l|}{c^{-4/3}}
  \;=\; c^{5/3}|k-l|.
\]
This is already a significant reduction from the raw Widom bound $c^{7/3}|k-l|$
(by a factor $c^{2/3}$), and it shows that the worst-case growth
originates in the transition regime.
The further reduction from $c^{5/3}|k-l|$ to the required $c^{1/2}$
(by an additional $c^{7/6}$ per unit $|k-l|$) is expected from
second-order transition asymptotics involving finer Airy-function estimates,
but is not currently established.
\end{remark}

\begin{assumption}[Prefactor bound, pointwise]\label{ass:B}
For all $k \neq l$ with $k, l \leq N \sim c^{1/3}$:
\[
  c \cdot \frac{|\lam{l}-\lam{k}|}{(1-\lam{k})(1-\lam{l})} \leq C_2 c^{1/2}.
\]
\end{assumption}

\begin{assumption}[Schur-summability hypothesis]\label{ass:Bprime}
The Schur row sums satisfy a uniform $c^{-1/2}$-improvement:
\[
  \sup_{i \leq N}\; \sum_{j \neq i,\, j \leq N}
  P_{ij} \cdot |I_{ij}|
  \leq C\, c^{-1/2} \log c.
\]
This is \emph{not} a consequence of Assumption~\ref{ass:B}.
It requires collective cancellation in the row sum beyond pointwise bounds;
see Section~\ref{sec:dyadic} for why this is the unique remaining obstacle.
\end{assumption}

\subsection{Matrix entry bound}

\begin{proposition}[Matrix entry bound]\label{prop:Tij}
Under Assumption~\ref{ass:B}, for $i \neq j$, $i, j \leq N$:
\[
  |T_{ij}| \leq \frac{C_1 C_2}{|i-j|}.
\]
The factors $c^{1/2}$ cancel; the bound is $c$-independent.
\end{proposition}

\begin{proof}
$|T_{ij}| \lesssim P_{ij} \cdot |I_{ij}|
  \leq C_2 c^{1/2} \cdot C_1 c^{-1/2}/|i-j| = C_1 C_2/|i-j|$.
\end{proof}

\begin{remark}[$c$-power balance]\label{rem:balance}
$c^{1/2}_{\rm Ass.\,B} \times c^{-1/2}_{\rm Lem.\,\ref{lem:RL}} = c^0$.
Conditional on Assumption~\ref{ass:B}, the norm grows at most like $\log c$
instead of $c^{8/3}$.
\end{remark}

\subsection{Proof of Theorem~\ref{thm:main}}

\begin{proof}[Proof of Theorem~\ref{thm:main}]
By~\eqref{eq:diagzero}: $T_{ii} = 0$.

\noindent\textbf{Part~(a).}
Proposition~\ref{prop:Tij} gives $|T_{ij}| \leq C_1 C_2/|i-j|$.
Schur test:
\[
  \sup_i \sum_{j \neq i} |T_{ij}|
  \leq C_1 C_2 H_N
  \leq C_1 C_2\bigl(1 + \tfrac{1}{3}\log c + \log A\bigr)
  =: C(A)\log c.
\]
Since $C(A)\log c \to \infty$, this is a norm bound only.

\noindent\textbf{Part~(b).}
Assumption~\ref{ass:Bprime} gives
$\sup_i \sum_{j\neq i} P_{ij}|I_{ij}| \leq Cc^{-1/2}\log c$,
so $\|D\mathcal{T}_c^{(N)}\| \leq Cc^{-1/2}\log c \to 0$.
\end{proof}

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{llll}
\hline
\textbf{Step} & \textbf{Bound} & \textbf{Source} & \textbf{Status} \\
\hline
Diagonal vanishing & $T_{ii} = 0$ & commutator & \checkmark{} unconditional \\
IBP integral & $|I_{ij}| \leq C_1 c^{-1/2}/|i-j|$ & Lemma~\ref{lem:RL}
  & \checkmark{} unconditional \\
Bulk barrier identified & boundary term at $s_*=1-c^{-2/3}$ dominates
  & Prop.~\ref{prop:bulk_barrier} & \checkmark{} unconditional \\
TZ Schur contribution & $o(1) \to 0$ & Lemma~\ref{lem:zone_schur}
  & \checkmark{} unconditional \\
Prefactor (pointwise) & $P_{ij} \leq C_2 c^{1/2}$ & Assumption~\ref{ass:B}
  & \textbf{open (OP~\ref{op:prefactor}(1))} \\
$c$-power cancellation & $c^{1/2} \times c^{-1/2} = O(1)$ & Remark~\ref{rem:balance}
  & conditional on Assumption~\ref{ass:B} \\
Schur sum & $\geq (\kappa/2)\log c$, $\leq C\log c$
  & Thms~\ref{thm:nogo},~\ref{thm:main}(a)
  & \textbf{optimal: norm bound, not contraction} \\
\hline
Schur-sum improvement & $\leq C c^{-1/2}\log c \to 0$ & Theorem~\ref{thm:main}(b)
  & \textbf{open: requires Ass.~B$'$ (cancellation)} \\
\hline
\end{tabular}
\caption{Logical chain for Theorem~\ref{thm:main}.
The Schur norm bound $\log c$ is both an upper and lower bound
(No-go Theorem~\ref{thm:nogo}); it is optimal within the pointwise-bound class.
Only part~(b) (conditional on Assumption~B$'$) gives a contraction.}
\label{tab:mechanisms}
\end{table}

\begin{corollary}[Conditional Galerkin fixed point]\label{cor:fixedpoint}
Conditional on Assumption~\ref{ass:Bprime}: for $c \geq c_0$ there exists
a unique $g^{(N)} \in X_c^{(N)}$ with $\mathcal{T}_c^{(N)}(g^{(N)}) = g^{(N)}$,
and Picard iteration converges at rate $\leq C c^{-1/2}\log c$ per step.
\end{corollary}

%% ============================================================
\section{Dyadic Decomposition and No-go Theorem}
\label{sec:dyadic}
%% ============================================================

\subsection{Dyadic block decomposition}

For a fixed row index $i$, partition the off-diagonal set into dyadic blocks:
\begin{equation}\label{eq:dyadic_blocks}
  \mathcal{B}_m := \{ j : 2^m \leq |i-j| < 2^{m+1} \},
  \qquad m = 0, 1, \ldots, M,
  \quad M := \lfloor \log_2 N \rfloor.
\end{equation}
Each block $\mathcal{B}_m$ has at most $2^{m+1}$ elements.

\begin{proposition}[Block-wise Schur bound]\label{prop:dyadic_block}
Under Assumption~\ref{ass:B}, each dyadic block contributes:
\begin{equation}\label{eq:block_bound}
  \sum_{j \in \mathcal{B}_m} |T_{ij}|
  \leq \frac{C_1 C_2}{2^m} \cdot |\mathcal{B}_m|
  \leq 2 C_1 C_2.
\end{equation}
Every block contributes $O(1)$, independently of $m$.
\end{proposition}

\begin{proof}
For $j \in \mathcal{B}_m$: $|i-j| \geq 2^m$, so $|T_{ij}| \leq C_1 C_2/2^m$.
Summing over $|\mathcal{B}_m| \leq 2^{m+1}$ terms gives~\eqref{eq:block_bound}.
\end{proof}

\begin{proposition}[Total Schur sum]\label{prop:dyadic_total}
Under Assumption~\ref{ass:B}:
\begin{equation}\label{eq:schur_dyadic}
  \sum_{j \neq i} |T_{ij}|
  \leq \sum_{m=0}^{M} 2 C_1 C_2
  = 2C_1 C_2(M+1)
  \leq 2C_1 C_2(\log_2 N + 1)
  \sim C(A)\log c.
\end{equation}
Each of the $\sim \log N$ dyadic scales contributes $O(1)$;
the total grows logarithmically with no cancellation.
\end{proposition}

\begin{remark}[Sharpness of the dyadic bound]
For the discrete Hilbert kernel $T_{ij} = \kappa/|i-j|$,
equality is attained in every block up to constants,
and $\sum_{j \neq i} |T_{ij}| = \kappa H_{N-1}$ exactly.
No rearrangement or grouping can reduce this sum
without exploiting signed cancellation between terms.
\end{remark}

\subsection{Proof of the No-go Theorem}

\begin{proof}[Proof of Theorem~\ref{thm:nogo}]
For any interior row $i$ (so distances $d = 1, \ldots, N-1$ are all realised):
\[
  \sup_i \sum_{j \neq i}|T_{ij}|
  \geq \kappa \sum_{d=1}^{N-1} \frac{1}{d}
  = \kappa H_{N-1}.
\]
For the operator norm: $\norm{T}_{\ell^2} \geq \norm{T}_F/\sqrt{N}$,
and summing $\sim N$ rows each contributing $\geq \kappa H_{N-1}/2$
gives $\norm{T} \geq (\kappa/2) H_{N-1}$.
For $N = \lfloor Ac^{1/3}\rfloor$: $H_{N-1} \geq \tfrac{1}{3}\log c + O(1)$.
\end{proof}

\subsection{Structural characterisation of Assumption~B$'$}

The dyadic analysis shows precisely what cancellation is needed.
Define the signed block sum $S_m(i) := \sum_{j \in \mathcal{B}_m} T_{ij}$.
For a contraction, the signed sums must satisfy:
\begin{equation}\label{eq:cancellation_needed}
  \sum_{m=0}^{M} |S_m(i)| \leq C c^{-1/2}\log c,
\end{equation}
rather than the absolute bound $\sum_m 2C_1C_2 = O(\log c)$.
Assumption~B$'$ is equivalent to requiring oscillatory cancellation
within each dyadic block such that $|S_m(i)| \ll 2C_1C_2$ uniformly.

\begin{remark}[Three routes to~\eqref{eq:cancellation_needed}]
\label{rem:routes}
\begin{enumerate}[nosep,label=(\roman*)]
  \item \textbf{Sign oscillation of $T_{ij}$ in $j$}:
        if $T_{ij}$ alternates sign with period $\sim 1$,
        then $|S_m(i)| = O(1/2^m)$ and $\sum_m |S_m(i)| = O(1)$.
        This would follow from near-antisymmetry of $P_{ij}$ under $j \mapsto j+1$,
        plausible from oscillatory PSWF products but not established.
  \item \textbf{Near-diagonal sparsification}:
        if $P_{ij} \leq C c^{-1/2}$ (stronger pointwise Assumption~B),
        then $|S_m(i)| \leq Cc^{-1/2} \cdot 2$ and
        $\sum_m |S_m(i)| \leq 2Cc^{-1/2}\log c$.
        Closing the gap from $c^{5/3}|k-l|$ to $c^{-1/2}$
        requires a reduction by $c^{13/6}$ (see Remark~\ref{rem:assB_heuristic})
        and is analytically very demanding.
  \item \textbf{Weighted block suppression}:
        a suitable $\ell^2$ weighted norm downweights far-diagonal blocks.
        Section~\ref{sec:spectral_weights} shows all natural candidates fail.
\end{enumerate}
\end{remark}

\begin{remark}[Heuristic phase structure of $T_{ij}$]\label{rem:phase_Tij}
Under a WKB approximation, products of PSWFs admit an oscillatory representation
\[
  \pswf{i}(s)\pswf{j}(s)
  \;\approx\;
  A_{ij}(s)\,\cos(\theta_i(s) - \theta_j(s)),
\]
with slowly varying amplitude $A_{ij}$ and rapidly varying phases $\theta_k$.
Formally inserting this into the definition of $T_{ij}$ suggests a kernel
\[
  T_{ij}
  \;\approx\;
  \frac{\widetilde{A}_{ij}}{|i-j|}
  \cos(\Theta_i - \Theta_j),
\]
for effective phases $\Theta_k$ arising from the $s$-integration.
If the phase differences $\Theta_i - \Theta_j$ exhibit sufficient
non-degeneracy or equidistribution in $j$ (for fixed $i$), then the dyadic
block sums $S_m(i)$ may exhibit cancellation analogous to oscillatory
singular integrals, yielding~\eqref{eq:cancellation_needed}.
This provides a Calder\'{o}n--Zygmund-type heuristic interpretation of
Assumption~B$'$: the improvement from $\log N$ to $o(\log N)$ must arise
from oscillatory cancellation, not pointwise decay.
Making this rigorous requires quantitative control of the phases $\theta_k$
in the transition and bulk regimes, which is currently open.
\end{remark}

\begin{conjecture}[Linear phase equidistribution]\label{conj:phase}
The effective WKB phases $\Theta_k^{(c)}$ of $\pswf{k}$ evaluated at
$s_* = 1 - c^{-2/3}$ satisfy, for $k \leq N \sim c^{1/3}$:
\[
  \Theta_k^{(c)} = \alpha^{(c)} k + \beta^{(c)} + O(k^2/c),
\]
where $\alpha^{(c)} \in (0, 2\pi)$ is bounded away from $0$ and $\pi$
uniformly in $c$.
Under this conjecture, Assumption~B$'$ follows from a standard
Dirichlet-kernel cancellation argument applied to the signed block sums
$S_m(i)$.
\end{conjecture}

\begin{remark}[Fourier-multiplier route to B$'$]\label{rem:fourier_route}
If $T_{ij} = f(i-j)/(i-j)$ for some bounded oscillatory $f : \mathbb{Z} \to [-1,1]$,
then $T$ is a discrete convolution operator and
\[
  \|T\|_{\ell^2 \to \ell^2}
  = \bigl\|\widehat{(f/|\cdot|)}\bigr\|_{\ell^\infty(\mathbb{Z})}
  = O(1),
\]
by $\ell^2$-boundedness of the discrete Hilbert transform with oscillatory weight.
This would establish B$'$ unconditionally, bypassing the Schur argument entirely.
The key hypothesis is \emph{approximate translation-invariance}:
$T_{ij} \approx g(i-j)$ for some $g$, i.e., $P_{ij} \cdot I_{ij}$ depends
primarily on $|i-j|$ rather than on $i$ and $j$ separately.
Whether the PSWF matrix $T$ has this structure is an open question
that would follow from Conjecture~\ref{conj:phase}.
\end{remark}

\subsection{Why spectral weights fail}
\label{sec:spectral_weights}

\begin{proposition}[Spectral weights are asymptotically constant]
\label{prop:spectral_weight_fail}
In the Galerkin window $k \leq N \sim c^{1/3}$:
$1 - \lambda_k \sim c^{-2/3}$ uniformly, so for any $\alpha > 0$:
\[
  w_k = (1-\lambda_k)^\alpha \sim c^{-2\alpha/3}
  \quad\text{uniformly in } k \leq N,
  \qquad
  \frac{w_j}{w_i} = 1 + O(c^{-1/3}).
\]
Spectral weighting does not modify the Schur kernel.
\textbf{Eigenvalue-based weighting cannot produce contraction.}
\end{proposition}

\begin{proof}
All indices lie in the transition regime where
$1 - \lambda_k \asymp c^{-2/3}$ uniformly (Widom transition asymptotics).
Hence $w_j/w_i \to 1$.
\end{proof}

\begin{remark}[Structural reason for the failure]
\label{rem:spectral_weight_structural}
The weight $w_k = (1-\lambda_k)^\alpha$ is blind to index structure:
in the transition window all eigenvalues lie on a flat plateau,
so the weight is essentially constant.
The Schur kernel $1/|i-j|$ is an index-space object;
an effective weight must distinguish indices, not eigenvalues.
\end{remark}

\begin{remark}[Index-based weights also fail]\label{rem:index_weights}
\begin{enumerate}[nosep,label=(\alph*)]
  \item \textbf{Power weight} $w_k = (1+k)^\beta$:
        For the central row $i \sim N/2$, the weighted row sum is
        $\sum_{d=1}^{N}(1+d/i)^\beta/d \sim \log N$ for all $\beta$.
  \item \textbf{Exponential weight} $w_k = e^{\gamma k/N}$:
        Weighted kernel $e^{\gamma d/N}/d$; row sum
        $\sum_{d=1}^{N} e^{\gamma d/N}/d \sim e^\gamma \log N$.
\end{enumerate}
Both fail because $1/d$ is not summable, and a weight $\phi(d)/d$ is summable
only if $\phi(d) \to 0$ (far-diagonal downweighting),
but such asymmetry cannot hold simultaneously for all rows.
\end{remark}

\begin{corollary}[No naive weight works]\label{cor:no_naive_weight}
No weight of spectral, power, or exponential type can make the weighted
Schur row sum $\sup_i \sum_j (w_j/w_i)/|i-j|$ asymptotically smaller than $\log N$.
\end{corollary}

\begin{openproblem}[Two-weight Schur estimate]\label{op:weightnorm_precise}
Find positive sequences $\{w_k\}$ (not of spectral, power, or exponential type)
such that
\[
  \sup_{i \leq N} \sum_{j \neq i} \frac{w_j}{w_i} \cdot \frac{C_1 C_2}{|i-j|}
  < 1.
\]
If no symmetric weight works, find $\{u_k\}$, $\{v_k\}$ such that
the two-weight Schur condition holds:
\[
  \sum_j v_j |T_{ij}| \leq u_i
  \quad\text{and}\quad
  \sum_j u_j^{-1} |T_{ij}| \leq v_i^{-1}
  \quad\text{for all } i.
\]
\end{openproblem}

%% ============================================================
\section{Open Problems}
\label{sec:open}
%% ============================================================

\begin{openproblem}[Prefactor reduction and Schur improvement]
\label{op:prefactor}
Two separate problems:
\begin{enumerate}[nosep]
  \item Prove Assumption~\ref{ass:B} unconditionally:
        $P_{kl} \leq C_2 c^{1/2}$ for $k,l \leq N \sim c^{1/3}$.
        Remark~\ref{rem:assB_heuristic} reduces the gap from $c^{7/3}$
        to $c^{5/3}$ via transition-regime asymptotics;
        closing the remaining factor $c^{7/6}$ requires second-order
        Airy-function estimates.
  \item Prove Assumption~\ref{ass:Bprime} unconditionally.
        By Section~\ref{sec:dyadic}, this is equivalent to signed
        cancellation in dyadic row-sum blocks.
        Conjecture~\ref{conj:phase} (linear phase equidistribution)
        would imply B$'$ via a Dirichlet-kernel argument,
        and is the most structurally natural route.
\end{enumerate}
\end{openproblem}

\begin{openproblem}[Weighted norm approach]\label{op:weightnorm}
Find a $c$-dependent weighted norm $\|\cdot\|_w$ on $X_c^{(N)}$
under which $\|D\mathcal{T}_c^{(N)}\|_w < 1$ unconditionally.
By Proposition~\ref{prop:spectral_weight_fail} and
Corollary~\ref{cor:no_naive_weight}, spectral, power, and exponential
index weights all fail.
The correct weight, if it exists, must be asymmetric or of two-weight type
(Open Problem~\ref{op:weightnorm_precise}).
\end{openproblem}

\begin{openproblem}[Bulk--transition coupling and Airy--WKB matching]
\label{op:bulk_refined}
By Proposition~\ref{prop:bulk_barrier}, the improved bound
\[
  |I_{kl}^{\rm bulk}| \lesssim \frac{c^{-2/3}}{|k-l|}
\]
is equivalent to cancellation of the boundary term at $s_* = 1-c^{-2/3}$.

\medskip
\noindent\textbf{Approach via Airy coordinates.}
Set $s = 1 - c^{-2/3}x$, $x = O(1)$.
In these coordinates:
\begin{itemize}[nosep]
  \item $\pswf{k}$ transitions from $\Ai(c^{2/3}\zeta_k)$ (monotone, $x \gg 1$)
        to oscillatory WKB ($x \ll 1$).
  \item The difference phase linearises:
        $\theta_k(s) - \theta_l(s) \approx (k-l)\phi(x)$
        for some explicit $\phi$.
  \item The boundary term becomes an explicit integral in $x$
        involving $\Ai \cdot \Ai$ products.
\end{itemize}
If this integral exhibits oscillatory cancellation (Airy stationary phase),
the boundary term reduces to $O(c^{-2/3}/|k-l|)$.
If established, Theorem~\ref{thm:main}(a) upgrades unconditionally to
$\|D\mathcal{T}_c^{(N)}\| \lesssim c^{-1/6}\log c \to 0$,
eliminating the need for Assumption~B$'$.
\end{openproblem}

\begin{openproblem}[Optimal Galerkin dimension]\label{op:Nopt}
Determine the maximal $N(c)$ for which the Schur bound is below~$1$.
Conjecture: $N_{\rm opt} \sim c^{1/2}$.
\end{openproblem}

\begin{openproblem}[Self-adjointness and spectral norm]\label{op:Gershgorin}
Establish approximate self-adjointness $T_{ij} \approx \overline{T_{ji}}$
up to $O(c^{-1})$, to sharpen the Schur bound to a spectral radius bound.
\end{openproblem}

%% ============================================================
\section{Relation to Assumption A}
%% ============================================================

\begin{itemize}[nosep]
  \item \textbf{Unconditionally proved:}
        $|I_{kl}| \leq C_1 c^{-1/2}/|k-l|$ (Lemma~\ref{lem:RL});
        $T_{ii} = 0$; $\|DT\|_{L^2} \gtrsim c^{8/3}$;
        $\log c$ growth is optimal in the pointwise-bound class
        (Theorem~\ref{thm:nogo});
        spectral and naive index weights do not improve the Schur kernel
        (Proposition~\ref{prop:spectral_weight_fail},
        Corollary~\ref{cor:no_naive_weight});
        transition-zone Schur contribution is $o(1)$, bulk dominates
        (Lemma~\ref{lem:zone_schur});
        bulk oscillation barrier precisely identified at $s_* = 1-c^{-2/3}$
        (Proposition~\ref{prop:bulk_barrier}).

  \item \textbf{Under Assumption~B:}
        $|T_{ij}| \leq C_1 C_2/|i-j|$;
        $\|D\mathcal{T}_c^{(N)}\| \leq C\log c$ (norm estimate only).

  \item \textbf{Under Assumptions~B and~B$'$:}
        $\|D\mathcal{T}_c^{(N)}\| \leq Cc^{-1/2}\log c$;
        contraction and conditional Galerkin fixed point
        (Corollary~\ref{cor:fixedpoint}).

  \item \textbf{Three open routes, in increasing depth:}
        \begin{enumerate}[nosep,label=(\roman*)]
          \item Airy--WKB matching at $s_*$
                (Open Problem~\ref{op:bulk_refined}):
                unconditional contraction under Assumption~B alone,
                rate $c^{-1/6}\log c$.
          \item Linear phase equidistribution
                (Conjecture~\ref{conj:phase} $\Rightarrow$ Assumption~B$'$
                via Dirichlet-kernel / Fourier-multiplier argument,
                Remark~\ref{rem:fourier_route}):
                contraction under Assumption~B.
          \item Two-weight Schur estimate
                (Open Problem~\ref{op:weightnorm}):
                contraction without Assumption~B$'$, no Schur required.
        \end{enumerate}
        Route~(i) is the most immediate target:
        it reduces to a concrete computation in Airy coordinates at
        $s_* = 1 - c^{-2/3}$.
\end{itemize}

%% ============================================================
\section*{Acknowledgements}
%% ============================================================

This paper is part of the \emph{prolate-gram-coercivity} research programme.
The dyadic decomposition, No-go Theorem, structural characterisation
of Assumption~B$'$ via signed block cancellation, the proof that
spectral and index weights cannot improve the Schur kernel,
the identification of the bulk oscillation barrier
(Proposition~\ref{prop:bulk_barrier}), and the Fourier-multiplier
interpretation of Assumption~B$'$ (Remark~\ref{rem:fourier_route},
Conjecture~\ref{conj:phase}) were developed in working sessions, May 2026.

%% ============================================================
\begin{thebibliography}{99}

\bibitem{paper13cc}
  \textit{Companion paper on RHP asymptotics and Assumption~A},
  prolate-gram-coercivity programme, internal manuscript, 2025--2026.

\bibitem{Widom64}
  H.~Widom,
  \textit{Asymptotic behavior of the eigenvalues of certain integral
  equations~II},
  Arch.\ Rational Mech.\ Anal.\ \textbf{17} (1964), 215--229.

\bibitem{HMW73}
  R.~Hunt, B.~Muckenhoupt, R.~Wheeden,
  \textit{Weighted norm inequalities for the conjugate function and
  Hilbert transform},
  Trans.\ Amer.\ Math.\ Soc.\ \textbf{176} (1973), 227--251.

\bibitem{CF74}
  R.~R.~Coifman, C.~Fefferman,
  \textit{Weighted norm inequalities for maximal functions and singular
  integrals},
  Studia Math.\ \textbf{51} (1974), 241--250.

\bibitem{Kato66}
  T.~Kato,
  \textit{Perturbation Theory for Linear Operators},
  Springer, 1966.

\bibitem{SP61}
  D.~Slepian, H.~O.~Pollak,
  \textit{Prolate spheroidal wave functions, Fourier analysis and
  uncertainty~I},
  Bell System Tech.\ J.\ \textbf{40} (1961), 43--63.

\bibitem{OsipovRokhlinXiao2013}
  A.~Osipov, V.~Rokhlin, H.~Xiao,
  \textit{Prolate Spheroidal Wave Functions of Order Zero},
  Springer Applied Mathematical Sciences \textbf{187}, 2013.

\bibitem{Olver1974}
  F.~W.~J.~Olver,
  \textit{Asymptotics and Special Functions},
  Academic Press, 1974.

\bibitem{GarciaTyler2020}
  S.~R.~Garcia, R.~A.~Horn,
  \textit{A Second Course in Linear Algebra},
  Cambridge University Press, 2017.

\end{thebibliography}

\end{document}
