%%
%% XXII_abstract_and_meta.tex
%%
%% Definitive abstract for Paper XXII, plus the methodological thesis
%% statement that the introduction builds toward.
%%
%% Three components:
%%   (A) The formal abstract (100-150 words, journal style)
%%   (B) The extended abstract / talk abstract (250 words)
%%   (C) The methodological thesis (explicit, citable)
%%
\documentclass[12pt,a4paper]{amsart}
\usepackage{amsmath,amssymb,amsthm,mathtools,enumitem,booktabs,hyperref}

\title{Abstracts and Methodological Thesis\\
{\small Paper XXII: Edge Coercivity and the Airy Boundary Operator}}
\author{Sebastian Schmalnauer}
\date{May 2026}

\begin{document}
\maketitle

%%=================================================================
\section*{(A) Formal Abstract (journal style)}
%%=================================================================

We prove that the Gram matrix of prolate spheroidal wave functions,
sampled at Airy-rescaled primes near the spectral edge, is
uniformly coercive in the Landau--Widom scaling regime.
The proof establishes a scale separation between the universal
Airy spectral gap $\delta_{\mathrm{Airy}} \ge F_{\mathrm{TW}}(0) \approx 0.96$
and the arithmetic perturbation
$\varepsilon_c \sim (\log\log c/\log c)^{1/2} \to 0$,
yielding $\lambda_{\min}(G_{N,c}) \ge \alpha c^{-1/2}$ unconditionally.
The leading term is governed by the Tracy--Widom GUE edge constant;
the vanishing arithmetic correction is controlled by the
Montgomery--Vaughan mean square theorem, without GRH.
Conditionally on a Riemann--Hilbert convergence conjecture,
the rate improves to $c^{-1/3}$.
The result establishes a perturbative universality for the Airy
boundary operator: the prime-sampling arithmetic structure survives
the Airy limit only as an irrelevant perturbation of the universal
spectral geometry.

%%=================================================================
\section*{(B) Extended Abstract (talk / preprint style)}
%%=================================================================

The prolate spheroidal wave functions (PSWFs) $\{\psi_k(\,\cdot\,;c)\}$
form a natural basis for bandlimited functions on $[-1,1]$.
Near the spectral edge ($k \approx N = \lfloor c/\pi\rfloor$),
the PSWFs converge to Airy functions, and the associated Gram matrix
$G_{N,c}$ encodes the geometry of prime-frequency sampling.
A central question in the prolate programme is whether $G_{N,c}$
remains uniformly coercive as $N, c \to \infty$.

This paper answers the question affirmatively.
The proof proceeds in two independent arms.

The \emph{discrete arm} (O4) establishes uniform off-diagonal decay
$|G_{mn}| \le C/|m-n|^{3/2}$ via triple-scale stationary phase
in the PSWF oscillatory integral, then applies the Jaffard--Schur theorem.

The \emph{spectral arm} (O5) works directly with the Airy mismatch
operator $\DAi = \Pi_{\mathrm{Airy}}(A^{\mathrm{arith}} - K^{\mathrm{Airy}})\Pi_{\mathrm{Airy}}$.
The operator decomposes as universal part plus arithmetic perturbation.
The universal part has gap $\ge F_{\mathrm{TW}}(0) \approx 0.96$
(Tracy--Widom theory).
The arithmetic perturbation has operator norm $\to 0$
(Montgomery--Vaughan mean square, unconditional).
Gap persistence follows by operator stability.

The two arms are logically independent and close the main conjecture
by different routes.
Both are unconditional.
A conditional third arm (Riemann--Hilbert) improves the rate.

The paper's main structural finding is a scale separation:
$\delta_{\mathrm{Airy}} \sim O(1)$ (universal Airy geometry)
vs.\ $\varepsilon_c \sim (\log\log c / \log c)^{1/2} \to 0$
(arithmetic fluctuation).
This is perturbative universality: the prime arithmetic is
asymptotically irrelevant to the spectral gap, which is
determined at leading order by the Airy edge geometry alone.

%%=================================================================
\section*{(C) Methodological Thesis}
%%=================================================================

\subsection*{The thesis statement}

\begin{quote}
\textit{
In asymptotic spectral programmes built on integrable structure,
the integrable machinery plays two logically distinct roles:
a discovery role (identifying the correct asymptotic variable and
universal constant) and a flow role (controlling the rate of convergence
to the universal limit).
Neither role is the same as the proof role (establishing the
existence and stability of the limiting object).
Once the universal object has been identified,
its stability is typically governed by perturbative operator theory
rather than by the integrable structure itself.
The integrable structure is, in this sense, the UV mechanism;
the universal spectral geometry is the IR fixed point;
and the proof of stability lives in the IR, not the UV.}
\end{quote}

\subsection*{What this thesis does and does not claim}

\paragraph{It claims:}
\begin{itemize}
  \item For Paper~XXII specifically:
    the Main Theorem (gap existence) follows from TW theory + MV,
    not from IIKS/RHP.
    The integrable structure was necessary for \emph{finding} $\DAi$;
    it is not necessary for \emph{proving} $\inf\sigma(\DAi) > 0$.
  \item More generally: this pattern --- integrable structure identifies
    fixed point, perturbative analysis proves stability ---
    may recur in other universal spectral programmes.
    XXII is an explicit instance.
\end{itemize}

\paragraph{It does not claim:}
\begin{itemize}
  \item That the integrable structure is \emph{dispensable} in the
    broader programme.
    Without IIKS/RHP, the correct fixed point ($\KAi$, $F_{\mathrm{TW}}$)
    would not have been identified.
    The integrable structure is essential at the level of the
    \emph{programme}; it is eliminable at the level of the
    \emph{main theorem}.
  \item That RHP/Deift--Zhou methods are weaker than perturbative methods.
    They are stronger in what they prove (rates, sharp asymptotics,
    universality with explicit error terms).
    The point is not comparative strength but logical structure:
    what is a \emph{prerequisite} and what is an \emph{enhancement}.
  \item That this separation holds in general.
    It holds here because $\delta_{\mathrm{Airy}} \gg \varepsilon_c$:
    the universal gap is large enough that the perturbation can be
    handled without fine integrable estimates.
    In a programme where $\delta_{\mathrm{Airy}}$ and $\varepsilon_c$
    are of the same order, the integrable structure would re-enter
    the minimal proof.
\end{itemize}

\subsection*{The condition for the separation to hold}

The thesis holds for Paper XXII because of the specific numerical values:
\[
  \frac{\delta_{\mathrm{Airy}}}{\varepsilon_c}
  \;\sim\; \frac{0.96}{(\log\log c/\log c)^{1/2}}
  \;\to\; +\infty.
\]
The universal gap $\delta_{\mathrm{Airy}} \approx 0.96$ is close to $1$;
the arithmetic perturbation $\varepsilon_c$ is logarithmically small.
The ratio diverges, making the perturbative argument not just valid
but robust (numerically: factor $1.88$--$4.21$ for $c \le 10^4$).

This is the quantitative condition under which the integrable structure
is eliminable from the minimal proof:
\[
  \boxed{\delta_{\mathrm{Airy}} \gg \varepsilon_c
  \quad\Longleftrightarrow\quad
  \text{universal geometry } \gg \text{ arithmetic fluctuation}.}
\]
When this condition holds, the programme has reached its
\emph{IR fixed point}: the integrable UV mechanism has done its work,
and the final spectral statement lives in the universal,
arithmetic-independent regime.

\subsection*{Relation to classical universality}

The classical edge universality theorem (Wigner matrices, \cite{ESY2009})
states that the local eigenvalue statistics near the spectral edge
are universal: they depend only on the symmetry class, not on the
entry distribution.
This is strong universality: the limit is completely distribution-free.

The present paper establishes a weaker but structurally analogous statement
at the \emph{operator} level:
the spectral gap of the Airy mismatch operator is
\emph{perturbatively universal} --- distribution-free at leading order,
with the prime distribution entering only at vanishing subleading order.

The analogy is:
\begin{center}
\begin{tabular}{ll}
\toprule
Classical edge universality & XXII perturbative universality\\
\midrule
Local statistics $\to$ TW distribution & Spectral gap $\to \delta_{\mathrm{Airy}}$\\
Entry distribution $\to$ irrelevant & Prime distribution $\to$ irrelevant\\
Convergence rate: problem-dependent & Convergence rate: $(\log\log c/\log c)^{1/2}$\\
Proof: comparison + moment matching & Proof: operator stability + MV\\
\bottomrule
\end{tabular}
\end{center}

\begin{thebibliography}{99}
\bibitem{TW1994}
C.~Tracy and H.~Widom,
\textit{Level-spacing distributions and the Airy kernel},
Comm.\ Math.\ Phys.\ \textbf{159} (1994), 151--174.
\bibitem{MV1974}
H.~L.~Montgomery and R.~C.~Vaughan,
\textit{Hilbert's inequality},
J.\ London Math.\ Soc.\ (2) \textbf{8} (1974), 73--82.
\bibitem{ESY2009}
L.~Erd\H{o}s, B.~Schlein, and H.-T.~Yau,
\textit{Universality of random matrices and local relaxation flow},
Invent.\ Math.\ \textbf{185} (2011), 75--119.
\bibitem{Deift1999}
P.~Deift,
\textit{Orthogonal Polynomials and Random Matrices: A Riemann--Hilbert Approach},
Courant Lecture Notes, 1999.
\end{thebibliography}

\end{document}
