%%
%% XXII_architectural_remark.tex
%%
%% The Renormalisation Interpretation of Paper XXII.
%% Architectural remark: IIKS/RHP as coordinate system, not proof core.
%%
%% This is a mathematical remark, not a philosophical note.
%% It makes precise three claims:
%%   (A) The minimal proof of P13 does not use IIKS/RHP.
%%   (B) The Airy geometry is the RG fixed point of the programme.
%%   (C) Papers XVIII-XXI were necessary to identify the right variable,
%%       but are not logically required in the final argument.
%%
\documentclass[12pt,a4paper]{amsart}
\usepackage{amsmath,amssymb,amsthm,mathtools,enumitem,booktabs,hyperref}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{observation}[theorem]{Observation}

\newcommand{\DAi}{D^{\mathrm{Airy}}}
\newcommand{\KAi}{K^{\mathrm{Airy}}}
\newcommand{\PAi}{\Pi_{\mathrm{Airy}}}
\newcommand{\Aarith}{A^{\mathrm{arith}}}
\newcommand{\norm}[1]{\|#1\|}

\title{Architectural Remark: IIKS/RHP as Coordinate System\\
{\normalsize Paper XXII --- The Renormalisation Interpretation}}
\author{Sebastian Schmalnauer}
\date{May 2026}

\begin{document}
\maketitle

\begin{abstract}
We make precise the observation that the minimal proof of the
Main Theorem of Paper~XXII (Theorem P13) does not use the
IIKS identity, the Riemann--Hilbert formalism, or the
Deift--Zhou steepest descent method.
These structures were essential for \emph{identifying} the correct
asymptotic variable (the Airy mismatch operator $\DAi$),
but are not logically required once this variable is identified.
We interpret this as an asymptotic renormalisation:
the Airy geometry is the \emph{stable fixed point} of the programme,
and all arithmetic and integrable-systems structure produces only
irrelevant perturbations in the renormalisation group sense.
\end{abstract}

%%=================================================================
\section{The Minimal Proof Skeleton}
\label{sec:skeleton}
%%=================================================================

\begin{observation}[Minimal proof of P13]\label{obs:minimal}
The Main Theorem (P13) follows from exactly four inputs:
\begin{enumerate}[\rm(I)]
  \item \textbf{Operator identification.}
    $\DAi = \PAi(\Aarith - \KAi)\PAi$.
    This is a definition (Paper XXII, Section~5), not a theorem.
  \item \textbf{Universal gap.}
    $\inf\sigma(\PAi(I - \KAi)\PAi) \ge F_{\mathrm{TW}}(0) \approx 0.96$.
    Source: Tracy--Widom 1994 + Bornemann 2010. No IIKS needed.
  \item \textbf{Arithmetic perturbation bound.}
    $\norm{\PAi(\Aarith - I)\PAi}_{\mathrm{op}} \le C\varepsilon_c \to 0$.
    Source: Montgomery--Vaughan 1974. No RHP needed.
  \item \textbf{Operator stability.}
    (I) + (II) + (III) $\Rightarrow$ $\inf\sigma(\DAi) > 0$.
    Source: min-max principle. Three lines.
\end{enumerate}
The proof is complete. No other input is required.
\end{observation}

\begin{remark}[What is NOT in the minimal proof]
The following structures appear in Papers~XII--XXI and in
the broader architecture of Paper~XXII, but are \emph{not used}
in the proof of Observation~\ref{obs:minimal}:
\begin{itemize}
  \item IIKS identity ($K_c(x,y) = F_c(x)^\top J G_c(y)/(x-y)$)
  \item Riemann--Hilbert Problem for $Y_c$
  \item Deift--Zhou steepest descent
  \item Assumption~A (existence of $g_c$)
  \item Assumption~B (steepest descent geometry)
  \item Conjecture~3.3 ($Y_c \to Y_{\mathrm{Airy}}$)
  \item The cubic Airy phase $\tfrac{1}{3}u^3$
  \item The Widom edge phase $\phi_c$
\end{itemize}
All of these are used only in the RHP arm (O1/O2/P8/P10),
which improves the \emph{rate} of the gap ($c^{-1/2} \to c^{-1/3}$)
but does not establish its \emph{existence}.
\end{remark}

%%=================================================================
\section{The Role of Papers XVIII--XXI}
\label{sec:role}
%%=================================================================

\begin{observation}[Discovery vs. proof]\label{obs:discovery}
The integrable-systems machinery (IIKS, RHP, Airy parametrix)
played two distinct roles in the development of Paper~XXII:
\begin{center}
\begin{tabular}{lll}
\toprule
Role & Mathematical content & Required in final proof?\\
\midrule
\textbf{Discovery} & identified $\DAi$ as the correct object & No\\
\textbf{Coordinates} & gave explicit formula for $K_c$ near the edge & No\\
\textbf{Rate refinement} & $\norm{K_c - \KAi}_{S_1} = O(c^{-1/3})$ & Conditionally (O1/O2)\\
\textbf{Context} & established Airy limit for PSWF (Papers XVIII--XX) & No\\
\bottomrule
\end{tabular}
\end{center}

\medskip
\noindent
Without Papers~XVIII--XXI, one would not know that $\DAi$ is the
relevant operator, nor that $F_{\mathrm{TW}}(0)$ is the right constant.
The integrable structure was \emph{epistemically necessary}:
it was needed to \emph{find} the right variable.

But once the variable $\DAi$ is identified, the integrable structure
can be \emph{logically removed} from the proof.
The operator $\DAi$ is defined independently of IIKS;
the bound $\inf\sigma \ge F_{\mathrm{TW}}(0) - C\varepsilon_c$ requires
only TW theory and Montgomery--Vaughan.

This is a clean separation between \emph{heuristic scaffolding}
and \emph{proof architecture}.
\end{observation}

%%=================================================================
\section{The Renormalisation Group Interpretation}
\label{sec:rg}
%%=================================================================

\begin{observation}[Airy geometry as RG fixed point]\label{obs:rg}
The structure of Theorem~P13 can be interpreted as a
renormalisation group (RG) fixed-point statement:
\begin{align*}
  &\text{\emph{Relevant direction:}}
    &\delta_{\mathrm{Airy}} &\sim O(1)
    &&\text{(universal Airy geometry --- does not flow)}\\
  &\text{\emph{Irrelevant direction:}}
    &\varepsilon_c &\sim \Bigl(\tfrac{\log\log c}{\log c}\Bigr)^{1/2} \to 0
    &&\text{(arithmetic fluctuation --- flows to zero)}
\end{align*}
The Airy geometry is the \emph{stable fixed point}:
as $c \to \infty$, the system flows toward the universal Airy operator,
and the arithmetic perturbation becomes irrelevant.

In the RG language:
\begin{itemize}
  \item $\KAi$ is the \emph{fixed-point operator}.
  \item $F_{\mathrm{TW}}(0)$ is the \emph{fixed-point value} of the universal gap.
  \item $\Aarith - I = \Ec$ is the \emph{irrelevant coupling}.
  \item The RG flow is $c \to \infty$; the flow equation is $\varepsilon_c \to 0$.
\end{itemize}
The condition for the gap to survive (O5) is exactly the condition
that the irrelevant coupling does not destabilise the fixed point:
$\varepsilon_c < \delta_{\mathrm{Airy}}$.
This is satisfied for all $c \ge c_0$ (Theorem P13).
\end{observation}

\begin{remark}[What flows and what does not]
The following quantities are \emph{relevant} (do not flow to zero
under $c \to \infty$) and determine the fixed point:
\begin{itemize}
  \item $F_{\mathrm{TW}}(0) \approx 0.9597$ (GUE edge universality constant)
  \item $\zeta(3/2) - 1 \approx 0.612$ (Jaffard coupling constant)
  \item $C_{\mathrm{crit}} = F_{\mathrm{TW}}(0)/(\zeta(3/2)-1) \approx 0.595$
    (threshold for the Jaffard route)
\end{itemize}
The following are \emph{irrelevant} (flow to zero):
\begin{itemize}
  \item $\varepsilon_c = \norm{\Ec}_{\mathrm{op}} \sim (\log\log c/\log c)^{1/2}$
  \item Off-diagonal fluctuations $|G_{mn} - G_{mn}^{\infty}|$ for $m \ne n$
  \item Finite-$c$ corrections to the Airy parametrix
\end{itemize}
The integrable-systems structure (IIKS, Widom phase, cubic phase)
appears only in the \emph{irrelevant} sector:
it controls corrections to the Airy limit, not the fixed point itself.
\end{remark}

%%=================================================================
\section{What O1/O2 Actually Adds}
\label{sec:o1o2}
%%=================================================================

\begin{observation}[The precise contribution of O1/O2]\label{obs:o1o2}
O1/O2 (the RHP arm: Assumptions A and B, Conjecture~3.3)
provide exactly one additional piece of information beyond the
minimal proof:
\begin{equation}\label{eq:rate}
  \norm{K_c - \KAi}_{S_1} = O(c^{-1/3})
  \quad\text{(conditional on Conjecture~3.3)}.
\end{equation}
This upgrades:
\begin{center}
\begin{tabular}{lll}
\toprule
Result & Gap scale & Source\\
\midrule
P13 (unconditional) & $\lambda_{\min}(G_{N,c}) \ge \alpha c^{-1/2}$ & O4 + BB3\\
P10 (conditional on~\eqref{eq:rate}) & $\lambda_{\min}(G_{N,c}) \ge \alpha c^{-1/3}$ & O1/O2 + P8\\
\bottomrule
\end{tabular}
\end{center}
The gap \emph{exists} (P13) without O1/O2.
The gap has a \emph{better rate} (P10) with O1/O2.

In the RG language:
O1/O2 characterise the \emph{speed of approach} to the fixed point,
not the fixed point itself.
They are a statement about the \emph{flow}, not the \emph{limit}.

This is why the DAG correctly identifies O1/O2 as off the minimal path:
they refine the programme, but do not drive it.
\end{observation}

%%=================================================================
\section{A Concise Summary}
\label{sec:summary}
%%=================================================================

\begin{observation}[The three-level structure]
Paper XXII has a clean three-level architecture:
\[
  \underbrace{\text{Level 1: Fixed point}}
    _{\delta_{\mathrm{Airy}} = 1 - \norm{\KAi}}
  \quad\longleftarrow\quad
  \underbrace{\text{Level 2: Perturbation}}
    _{\varepsilon_c = \norm{\Aarith - I}_{\mathrm{op}}}
  \quad\longleftarrow\quad
  \underbrace{\text{Level 3: Flow}}
    _{\norm{K_c - \KAi}_{S_1} = O(c^{-1/3})}
\]
\begin{itemize}
  \item \textbf{Level 1} (TW theory): establishes the fixed point.
    Used in: Theorem P13 (minimal proof).
  \item \textbf{Level 2} (Montgomery--Vaughan): bounds the perturbation.
    Used in: Theorem P13 (minimal proof).
  \item \textbf{Level 3} (IIKS/RHP/Deift--Zhou): characterises the flow.
    Used in: P10 (rate refinement, conditional on O1/O2).
\end{itemize}
The integrable systems machinery lives entirely at Level~3.
The gap lives at Level~1.
The separation is exact.
\end{observation}

\bigskip

\noindent
\textit{Note on the role of this document.}
This remark is not a philosophical digression.
It has a concrete function: it clarifies which parts of Paper~XXII
are logically load-bearing for the main theorem,
and which are load-bearing only for rate improvements or future work.
This distinction matters for the final write-up:
Level~1 and Level~2 must be fully proved;
Level~3 can be stated conditionally without blocking submission.

\end{document}
